---
title: Bioconductor notes
output:
  bookdown::html_document2:
    toc: true
    toc_float: true
fontsize: 11pt
bibliography: '`r system.file("REFERENCES.bib", package="pQTLtools")`'
csl: nature-genetics.csl
link-citations: yes
pkgdown:
  as_is: true
vignette: >
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteIndexEntry{Bioconductor Applications}
  %\VignetteEncoding{UTF-8}
---

```{r logo, echo = FALSE, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "bioconductor/"
)
```

This article collects notes on Bioconductor packages, made available here to faciliate their use and extensions.

```{r, message=FALSE, warning=FALSE}
pkgs <- c("AnnotationDbi", "Biostrings", "ComplexHeatmap", "DESeq2", "EnsDb.Hsapiens.v86",
          "FlowSorted.DLPFC.450k", "GeneNet", "GenomicFeatures", "RColorBrewer",
          "RMariaDB", "Rgraphviz", "S4Vectors", "SummarizedExperiment",
          "TxDb.Hsapiens.UCSC.hg38.knownGene", "bladderbatch", "clusterProfiler",
          "corpcor", "ensembldb", "fdrtool", "graph", "graphite", "heatmaply",
          "minfi", "org.Hs.eg.db", "plyr", "quantro", "recount3", "sva")
for (p in pkgs) if (length(grep(paste("^package:", p, "$", sep=""), search())) == 0) {
    if (!requireNamespace(p)) warning(paste0("This vignette needs package `", p, "'; please install"))
}
invisible(suppressMessages(lapply(pkgs, require, character.only = TRUE)))
```

# liftover

See `inst/turboman`in the source, <https://github.com/jinghuazhao/pQTLtools/tree/master/inst/turboman>, or `turboman/` directory in the installed package.

# Normalisation

## ComBat

This is the documentation example, based on Bioconductor 3.14.

```{r ComBat, fig.cap="ComBat example", messages=FALSE}
data(bladderdata)
edat <- bladderEset[1:50,]

pheno <- pData(edat)
batch <- pheno$batch
table(batch)
matboxplot(edat,batch,cex.axis=0.6,notch=TRUE,ylab="Expression")
matdensity(edat,batch,xlab=" ",ylab="density")
legend("topleft",legend=1:5,col=1:5,lty=1)

# 1. parametric adjustment
combat_edata1 <- ComBat(dat=edat, batch=batch, par.prior=TRUE, prior.plots=TRUE)

# 2. non-parametric adjustment, mean-only version
combat_edata2 <- ComBat(dat=edat, batch=batch, par.prior=FALSE, mean.only=TRUE)

# 3. reference-batch version, with covariates
mod <- model.matrix(~as.factor(cancer), data=pheno)
combat_edata3 <- ComBat(dat=edat, batch=batch, mod=mod, par.prior=TRUE, ref.batch=3, prior.plots=TRUE)
```

## quantro

This is also adapted from the package vignette but with `FlowSorted.DLPFC.450k` in place of `FlowSorted`.

```{r quantro, fig.cap="quantro example", messages=FALSE}
p <- getBeta(FlowSorted.DLPFC.450k,offset=100)
pd <- pData(FlowSorted.DLPFC.450k)
matboxplot(p, groupFactor = pd$CellType, xaxt = "n", main = "Beta Values")
matdensity(p, groupFactor = pd$CellType, xlab = " ", ylab = "density",
           main = "Beta Values", brewer.n = 8, brewer.name = "Dark2")
legend('top', c("NeuN_neg", "NeuN_pos"), col = c(1, 2), lty = 1, lwd = 3)
qtest <- quantro(object = p, groupFactor = pd$CellType)
if (FALSE)
{
  registerDoParallel(cores=10)
  qtestPerm <- quantro(p, groupFactor = pd$CellType, B = 1000)
  quantroPlot(qtestPerm)
}
```

# Differential expression

```{r DESeq2, fig.cap="DESeq2 example", messages=FALSE}
ex <- makeExampleDESeqDataSet(m=4)
dds <- DESeq(ex)
res <- results(dds, contrast=c("condition","B","A"))
rld <- rlogTransformation(ex, blind=TRUE)
dat <- plotPCA(rld, intgroup=c("condition"),returnData=TRUE)
percentVar <- round(100 * attr(dat,"percentVar"))
ggplot(dat, aes(PC1, PC2, color=condition, shape=condition)) +
geom_point(size=3) +
xlab(paste0("PC1:",percentVar[1],"% variance")) +
ylab(paste0("PC2:",percentVar[2],"% variance"))
ex$condition <- relevel(ex$condition, ref="B")
dds2 <- DESeq(dds)
res <- results(dds2)
knitr::kable(head(as.data.frame(res)))
```

See the package in action from a snakemake workflow @koster21.

# Gene co-expression and network analysis

A simple network is furnished with the `GeneNet` documentation example,

```{r GeneNet, fig.cap="GeneNet example", messages=FALSE}
## A random network with 40 nodes 
# it contains 780=40*39/2 edges of which 5 percent (=39) are non-zero
true.pcor <- ggm.simulate.pcor(40)
  
# A data set with 40 observations
m.sim <- ggm.simulate.data(40, true.pcor)

# A simple estimate of partial correlations
estimated.pcor <- cor2pcor( cor(m.sim) )

# A comparison of estimated and true values
sum((true.pcor-estimated.pcor)^2)

# A slightly better estimate ...
estimated.pcor.2 <- ggm.estimate.pcor(m.sim)
sum((true.pcor-estimated.pcor.2)^2)

## ecoli data 
data(ecoli)

# partial correlation matrix 
inferred.pcor <- ggm.estimate.pcor(ecoli)

# p-values, q-values and posterior probabilities for each potential edge 
test.results <- network.test.edges(inferred.pcor)

# best 20 edges (strongest correlation)
test.results[1:20,]

# network containing edges with prob > 0.9 (i.e. local fdr < 0.1)
net <- extract.network(test.results, cutoff.ggm=0.9)
net

# significant based on FDR cutoff Q=0.05?
num.significant.1 <- sum(test.results$qval <= 0.05)
test.results[1:num.significant.1,]

# significant based on "local fdr" cutoff (prob > 0.9)?
num.significant.2 <- sum(test.results$prob > 0.9)
test.results[test.results$prob > 0.9,]

# parameters of the mixture distribution used to compute p-values etc.
c <- fdrtool(sm2vec(inferred.pcor), statistic="correlation")
c$param

## A random network with 20 nodes and 10 percent (=19) edges
true.pcor <- ggm.simulate.pcor(20, 0.1)

# convert to edge list
test.results <- ggm.list.edges(true.pcor)[1:19,]
nlab <- LETTERS[1:20]

# graphviz
# network.make.dot(filename="test.dot", test.results, nlab, main = "A graph")
# system("fdp -T svg -o test.svg test.dot")

# Rgraphviz
gr <- network.make.graph( test.results, nlab)
gr
num.nodes(gr)
edge.info(gr)
gr2 <- network.make.graph( test.results, nlab, drop.singles=TRUE)
gr2
num.nodes(gr2)
edge.info(gr2)

# plot network
plot(gr, "fdp")
plot(gr2, "fdp")
```

A side-by-side heatmaps

```{r heatmap, fig.cap="Heatmaps", messages=FALSE}
set.seed(123454321)
m <- matrix(runif(2500),50)
r <- cor(m)
g <- as.matrix(r>=0.7)+0
f1 <- Heatmap(r)
f2 <- Heatmap(g)
f <- f1+f2
draw(f)

df <- normalize(mtcars)
heatmaply(df,k_col=5,k_row=5,colors = colorRampPalette(brewer.pal(3, "RdBu"))(256))
```

and a module analysis with WGCNA,

```r
pwr <- c(1:10, seq(from=12, to=30, by=2))
sft <- pickSoftThreshold(dat, powerVector=pwr, verbose=5)
ADJ <- abs(cor(dat, method="pearson", use="pairwise.complete.obs"))^6
dissADJ <- 1-ADJ
dissTOM <- TOMdist(ADJ)
TOM <- TOMsimilarityFromExpr(dat)
Tree <- hclust(as.dist(1-TOM), method="average")
for(j in pwr)
{
  pam_name <- paste0("pam",j)
  assign(pam_name, pam(as.dist(dissADJ),j))
  pamTOM_name <- paste0("pamTOM",j)
  assign(pamTOM_name,pam(as.dist(dissTOM),j))
  tc <- table(get(pam_name)$clustering,get(pamTOM_name)$clustering)
  print(tc)
  print(diag(tc))
}
colorStaticTOM <- as.character(cutreeStaticColor(Tree,cutHeight=.99,minSize=5))
colorDynamicTOM <- labels2colors(cutreeDynamic(Tree,method="tree",minClusterSize=5))
Colors <- data.frame(pamTOM6$clustering,colorStaticTOM,colorDynamicTOM)
plotDendroAndColors(Tree, Colors, dendroLabels=FALSE, hang=0.03, addGuide=TRUE, guideHang=0.05)
meg <- moduleEigengenes(dat, color=1:ncol(dat), softPower=6)
```

# Meta-data

This section is based on package `recount3`.

```r
hs <- available_projects()
dim(subset(hs,file_source=="gtex"))
annotation_options("human")
blood_rse <- create_rse(subset(hs,project=="BLOOD"))
metadata(blood_rse)
rowRanges(blood_rse)
colnames(colData(blood_rse))[1:20]
expand_sra_attributes(blood_rse)
```

# Pathway and enrichment analysis

```{r pathway-enrichment, fig.cap="enricher example", messages=FALSE}
reactome <- pathways("hsapiens", "reactome")
kegg <- pathways("hsapiens","kegg")
pharmgkb <- pathways("hsapiens","pharmgkb")
nodes(kegg[[21]])
kegg_t2g <- ldply(lapply(kegg, nodes), data.frame)
names(kegg_t2g) <- c("gs_name", "gene_symbol")
VEGF <- subset(kegg_t2g,gs_name=="VEGF signaling pathway")[[2]]
eKEGG <- enricher(gene=VEGF, TERM2GENE = kegg_t2g,
                  universe=,
                  pAdjustMethod = "BH",
                  pvalueCutoff = 0.1, qvalueCutoff = 0.05,
                  minGSSize = 10, maxGSSize = 500)

```

# Peptide sequence

Here is an example for PROC_HUMAN, which is handled by the `Biostrings` package,

```{r a1bg, fig.cap="peptide association plot",fig.height=6,fig.width=10,messages=FALSE}
fasta_file_path <- 'https://rest.uniprot.org/uniprotkb/P04070.fasta'
fasta_sequences <- Biostrings::readAAStringSet(fasta_file_path, format = "fasta")
AA_sequence <- fasta_sequences[[1]]
cat("Sequence:", toString(AA_sequence), "\n")
iso_442688365 <- 'TDGEGALSEPSATVTIEELAAPPPPVLMHHGESSQVLHPGNK'
match_position <- regexpr(iso_442688365, AA_sequence)
match_position
mp <- matchPattern(iso_442688365,AA_sequence)
mp
protein <- "PROC"
cistrans <- read.csv(paste0("~/pQTLtools/tests","/",protein,".cis.vs.trans"))
load(paste0("~/pQTLtools/tests/",protein,".rda"))
pQTLtools::peptideAssociationPlot(protein,cistrans)
```

# Transcript databases

An overview of annotation is available @carlson16.

```{r ts-data, messages=FALSE}
options(width=200)

# columns(org.Hs.eg.db)
# keyref <- keys(org.Hs.eg.db, keytype="ENTREZID")
# symbol_uniprot <- select(org.Hs.eg.db,keys=keyref,columns = c("SYMBOL","UNIPROT"))
# subset(symbol_uniprot,SYMBOL=="MC4R")

x <- EnsDb.Hsapiens.v86
listColumns(x, "protein", skip.keys=TRUE)
listGenebiotypes(x)
listTxbiotypes(x)
listTables(x)
metadata(x)
organism(x)
returnFilterColumns(x)
seqinfo(x)
seqlevels(x)
updateEnsDb(x)

genes(x, columns=c("gene_name"), filter=list(SeqNameFilter("X"), GeneBiotypeFilter("protein_coding")))
transcripts(x, columns=listColumns(x, "tx"), filter = AnnotationFilterList(), order.type = "asc", return.type = "GRanges")

txdbEnsemblGRCh38 <- makeTxDbFromEnsembl(organism="Homo sapiens", release=98)
txdb <- as.list(txdbEnsemblGRCh38)
lapply(txdb,head)

txdb <- TxDb.Hsapiens.UCSC.hg38.knownGene

# liverExprs <- quantifyExpressionsFromBWs(txdb = txdb,BWfiles=,experimentalDesign=)
```

# Bionconductor forum

Web: [https://support.bioconductor.org/](https://support.bioconductor.org/)

# Bioconductor/CRAN packages

Package | Description
--------|------------
**Bioconductor** |
AnnotationDbi | AnnotationDb objects and their progeny, methods etc.
Biobase | Base functions for Bioconductor
Biostrings | Efficient manipulation of biological strings
org.Hs.eg.db | Conversion of Entrez ID -- gene symbols
EnsDb.Hsapiens.v86 | Exposes an annotation databases generated from Ensembl
ensembldb | Retrieve annotation data from an Ensembl based package
TxDb.Hsapiens.UCSC.hg38.knownGene | Annotation of the human genome
INSPEcT | Quantification of the intronic and exonic gene features and the post-transcriptional regulation analysis
graphite | GRAPH Interaction from pathway topological environment
clusterProfiler | Functional profiles for genes and gene clusters
DESSeq2 | Differential gene expression analysis based on the negative binomial distribution
edgeR | Empirical analysis of digital gene expression
WGCNA | Weighted correlation network analysis
ComplexHeatmap | Make complex heatmaps
recount3 | Interface to uniformly processed RNA-seq data
Pi | Priority index, leveraging genetic evidence to prioritise drug targets at the gene and pathway level
quantro | A test for when to use quantile normalisation
FlowSorted.DLPFC.450k | Illumina HumanMethylation data on sorted frontal cortex cell populations
Rgraphiz | Interfaces R with the AT&T graphviz library for plotting R graph objects from the graph package
sva | Surrogate Variable Analysis
doParallel | Foreach Parallel Adaptor for the 'parallel' Package
**CRAN** |
GeneNet | Modeling and Inferring Gene Networks
RColorBrewer | ColorBrewer Palettes
ggplot2 | Data Visualisations Using the grammar of graphics
heatmaply | Interactive Cluster Heat Maps Using `plotly` and `ggplot2`
pheatmap | results visualisation
plyr | Splitting, applying and combining data

# References
