---
title: "ExpressionSet/SummarizedExperiment usage"
output:
  bookdown::html_document2:
    toc: true
    toc_float: true
fontsize: 11pt
bibliography: '`r system.file("REFERENCES.bib", package="pQTLtools")`'
csl: nature-genetics.csl
pkgdown:
  as_is: true
vignette: >
  %\VignetteIndexEntry{ExpressionSet usage}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
set.seed(0)
knitr::opts_chunk$set(
  out.extra = 'style="display:block; margin: auto"',
  fig.align = "center",
  fig.path = "esse/",
  collapse = TRUE,
  comment = "#>",
  dev = "png")
```

```{r, message=FALSE, warning=FALSE}
es_pkgs <- c("Biobase", "arrayQualityMetrics", "dplyr", "knitr", "mclust", "pQTLtools", "rgl", "scatterplot3d")
se_pkgs <- c("BiocGenerics","GenomicRanges","IRanges","MsCoreUtils","S4Vectors","SummarizedExperiment")
pkgs <- c(es_pkgs,se_pkgs)
for (p in pkgs) if (length(grep(paste("^package:", p, "$", sep=""), search())) == 0) {
    if (!requireNamespace(p)) warning(paste0("This vignette needs package `", p, "'; please install"))
}
invisible(suppressMessages(lapply(pkgs, require, character.only = TRUE)))
```

We start with Bioconductor/Biobase's ExpressionSet example and finish with a real application.

```{r, results='hide'}
dataDirectory <- system.file("extdata", package="Biobase")
exprsFile <- file.path(dataDirectory, "exprsData.txt")
exprs <- as.matrix(read.table(exprsFile, header=TRUE, sep="\t", row.names=1, as.is=TRUE))
pDataFile <- file.path(dataDirectory, "pData.txt")
pData <- read.table(pDataFile, row.names=1, header=TRUE, sep="\t")
all(rownames(pData)==colnames(exprs))
metadata <- data.frame(labelDescription=c("Patient gender",
                                          "Case/control status",
                                          "Tumor progress on XYZ scale"),
                       row.names=c("gender", "type", "score"))
phenoData <- Biobase::AnnotatedDataFrame(data=pData, varMetadata=metadata)
experimentData <- Biobase::MIAME(name="Pierre Fermat",
                                 lab="Francis Galton Lab",
                                 contact="pfermat@lab.not.exist",
                                 title="Smoking-Cancer Experiment",
                                 abstract="An example ExpressionSet",
                                 url="www.lab.not.exist",
                                 other=list(notes="Created from text files"))
exampleSet <- pQTLtools::make_ExpressionSet(exprs,phenoData,experimentData=experimentData,
                                            annotation="hgu95av2")
data(sample.ExpressionSet, package="Biobase")
identical(exampleSet,sample.ExpressionSet)
```

# data.frame

The great benefit is to use the object directly as a data.frame.

```{r}
lm.result <- Biobase::esApply(exampleSet,1,function(x) lm(score~gender+x))
beta.x <- unlist(lapply(lapply(lm.result,coef),"[",3))
beta.x[1]
lm(score~gender+AFFX.MurIL2_at,data=exampleSet)
```

# Composite plots

We wish to examine the distribution of each feature via histogram, scatter and boxplot.
One could resort to `esApply()` for its simplicity as before

```r
invisible(Biobase::esApply(exampleSet[1:2],1,function(x)
                           {par(mfrow=c(3,1));boxplot(x);hist(x);plot(x)}
))
```
but it is nicer to add feature name in the title.

```{r plots, fig.cap="Histogram, scatter & boxplots", fig.height=3}
par(mfrow=c(1,3))
f <- Biobase::featureNames(exampleSet[1:2])
invisible(sapply(f,function(x) {
                     d <- t(Biobase::exprs(exampleSet[x]))
                     fn <- Biobase::featureNames(exampleSet[x])
                     hist(d,main="",xlab=fn); plot(d, ylab=fn); boxplot(d,ylab=fn)
                   }
          )
)
```
where the expression set is indexed using feature name(s).

# Outlier detections

This illustrates one mechanism,

```r
list_outliers <- function(es, method="upperquartile")
                 arrayQualityMetrics::outliers(exprs(es),method=method)
for (method in c("KS","sum","upperquartile"))
{
  ZWK_outliers <- list_outliers(protein_ZWK,method=method)
  print(ZWK_outliers@statistic[ZWK_outliers@which])
}
```

# Clustering

We employ model-based clustering absed on principal compoents to see potential groupings in the data,

```{r mcpca, fig.cap="Three-group Clustering", fig.height=7, fig.width=7}
  pca <- prcomp(na.omit(t(Biobase::exprs(exampleSet))), rank=10, scale=TRUE)
  pc1pc2pc3 <- with(pca,x)[,1:3]
  mc <- mclust::Mclust(pc1pc2pc3,G=3)
  with(mc, {
      cols <- c("blue","red", "purple")
      s3d <- scatterplot3d::scatterplot3d(with(pca,x[,c(2,1,3)]),
                                          color=cols[classification],
                                          pch=16,
                                          type="h",
                                          main="Plot of the PC1, PC2 and PC3")
      s3d.coords <- s3d$xyz.convert(with(pca,x[,c(2,1,3)]))
      text(s3d.coords$x, 
           s3d.coords$y,   
           cex = 1.2,
           col = cols[classification],
           labels = row.names(pc1pc2pc3),
           pos = 4)
      legend("right", legend=levels(as.factor(classification)), col=cols[classification], pch=16)
      rgl::open3d(width = 500, height = 500)
      rgl::plot3d(with(pca,x[,c(2,1,3)]),cex=1.2,col=cols[classification],size=5)
      rgl::text3d(with(pca,x[,c(2,1,3)]),cex=1.2,col=cols[classification],texts=row.names(pc1pc2pc3))
      htmlwidgets::saveWidget(rgl::rglwidget(), file = "mcpca3d.html")
  })
```

An interactive version is also available,

```{r rgl3d, echo=FALSE}
htmltools::tags$iframe(src = "mcpca3d.html", width = "100%", height = "450px")
```

# Data transformation

Suppose we wish use log2 for those greater than zero but set those negative values to be missing.

```r
log2.na <- function(x) log2(ifelse(x>0, x, NA))
Biobase::exprs(exampleSet) <- log2.na(Biobase::exprs(exampleSet))
```

# Limit of detection (LOD)

We generate a `lod.max` ~ U[0,1] variable to experiment

```{r lod, fig.cap="LOD based on a random cutoff", fig.height=7}
Biobase::fData(exampleSet)
Biobase::fData(exampleSet)$lod.max <- apply(Biobase::exprs(exampleSet),1,quantile,runif(nrow(exampleSet)))
lod <- pQTLtools::get.prop.below.LLOD(exampleSet)
x <- dplyr::arrange(Biobase::fData(lod),desc(pc.belowLOD.new))
knitr::kable(head(lod))
plot(x[,2], main="Random quantile cutoff", ylab="<lod%")
```

The quantity has been shown to have a big impact on protein abundance and therefore pQTL detection as is shown with a real example.

```r
rm(list=ls())
dir <- "~/rds/post_qc_data/interval/phenotype/olink_proteomics/post-qc/"
eset <- readRDS(paste0(dir,"eset.inf1.flag.out.outlier.in.rds"))
x <- pQTLtools::get.prop.below.LLOD(eset)
annot <- Biobase::fData(x)
annot$MissDataProp <- as.numeric(gsub("\\%$", "", annot$MissDataProp))
plot(annot$MissDataProp, annot$pc.belowLOD.new, xlab="% <LLOD in Original",
     ylab="% <LLOD in post QC dataset", pch=19)
INF <- Sys.getenv("INF")
np <- read.table(paste(INF, "work", "INF1.merge.nosig", sep="/"), header=FALSE,
                 col.names = c("prot", "uniprot"))
kable(np, caption="Proteins with no pQTL")
annot$pQTL <- rep(NA, nrow(annot))
no.pQTL.ind <- which(annot$uniprot.id %in% np$uniprot)
annot$pQTL[no.pQTL.ind] <- "red"
annot$pQTL[-no.pQTL.ind] <- "blue"
annot <- annot[order(annot$pc.belowLOD.new, decreasing = TRUE),]
annot <- annot[-grep("^BDNF$", annot$ID),]
saveRDS(annot,file=file.path("~","pQTLtools","tests","annot.RDS"))
```

```{r test, fig.cap="LOD in SCALLOP-INF/INTERVAL", fig.height=8, fig.width=15}
annot <- readRDS(file.path(find.package("pQTLtools"),"tests","annot.RDS")) %>%
         dplyr::left_join(pQTLdata::inf1[c("prot","target.short","alt_name")],by=c("ID"="prot")) %>%
         dplyr::mutate(prot=if_else(is.na(alt_name),target.short,alt_name),order=1:n()) %>%
         dplyr::arrange(desc(order))
xtick <- seq(1, nrow(annot))
attach(annot)
par(mar=c(10,5,1,1))
plot(100-pc.belowLOD.new,cex=2,pch=19,col=pQTL,xaxt="n",xlab="",ylab="",cex.axis=0.8)
text(66,16,"IL-17C",offset=0,pos=2,cex=1.5,font=2,srt=0)
arrows(67,16,71,16,lwd=2)
axis(1, at=xtick, labels=prot, lwd.tick=0.5, lwd=0, las=2, hadj=1, cex.axis=0.8)
mtext("% samples above LLOD",side=2,line=2.5,cex=1.2)
mtext("Ordered protein",side=1,line=6.5,cex=1.2,font=1)
legend(x=1,y=25,c("without pQTL","with pQTL"),box.lwd=0,cex=2,col=c("red","blue"),pch=19)
detach(annot)
options(width=120)
knitr::kable(annot,caption="Summary statistics",row.names=FALSE)
```

# maEndtoEnd

Web: <https://bioconductor.org/packages/release/workflows/html/maEndToEnd.html>.

Examples can be found on PCA, heatmap, normalisation, linear models, and enrichment analysis from this Bioconductor package.

# SummarizedExperiment

This is a more modern construct. The following script is adapted from the documentation example, showing ranged summarized experiment (rse) adding missing values.

```{r, message=FALSE, warning=FALSE}
set.seed(123)
nrows <- 20
ncols <- 4
counts <- matrix(runif(nrows * ncols, 1, 1e4), nrows)
missing_indices <- sample(length(counts), size = 5, replace = FALSE)
counts[missing_indices] <- NA
rowRanges <- GenomicRanges::GRanges(rep(c("chr1", "chr2"), c(1, 3) * nrows / 4),
                            IRanges::IRanges(floor(runif(nrows, 1e5, 1e6)), width=ncols),
                            strand=sample(c("+", "-"), nrows, TRUE),
                            feature_id=sprintf("ID%03d", 1:nrows))
colData <- S4Vectors::DataFrame(Treatment=rep(c("ChIP", "Input"), ncols/2),
                                row.names=LETTERS[1:ncols])
rse <- SummarizedExperiment(assays=S4Vectors::SimpleList(counts=counts),
                            rowRanges=rowRanges, colData=colData)
assay(rse)
imputed <- MsCoreUtils::impute_knn(as.matrix(SummarizedExperiment::assay(rse)),2)
imputed_counts <- MsCoreUtils::impute_RF(as.matrix(assay(rse)),2)
imputed-imputed_counts
SummarizedExperiment::assays(rse) <- S4Vectors::SimpleList(counts=imputed_counts)
SummarizedExperiment::assay(rse)
SummarizedExperiment::assays(rse) <- S4Vectors::endoapply(SummarizedExperiment::assays(rse), asinh)
SummarizedExperiment::assay(rse)

SummarizedExperiment::rowRanges(rse)
SummarizedExperiment::rowData(rse)
SummarizedExperiment::colData(rse)

rse[, rse$Treatment == "ChIP"]

# the same ranges but different samples:
rse1 <- rse
rse2 <- rse1[, 1:3]
colnames(rse2) <- letters[1:ncol(rse2)]
cmb1 <- cbind(rse1, rse2)
dim(cmb1)
dimnames(cmb1)

# the same samples but different ranges:
rse1 <- rse
rse2 <- rse1[1:nrows, ]
rownames(rse2) <- letters[1:nrow(rse2)]
cmb2 <- rbind(rse1, rse2)
dim(cmb2)
dimnames(cmb2)

se0 <- as(rse, "SummarizedExperiment")
se0
as(se0, "RangedSummarizedExperiment")

# rowRanges(a SummarizedExperiment object) --> RangedSummarizedExperiment object:
se <- se0
rowRanges(se) <- rowRanges
se

stopifnot(identical(SummarizedExperiment::assays(se0), SummarizedExperiment::assays(rse)))
stopifnot(identical(dim(se0), dim(rse)))
stopifnot(identical(dimnames(se0), dimnames(rse)))
stopifnot(identical(SummarizedExperiment::rowData(se0), SummarizedExperiment::rowData(rse)))
stopifnot(identical(SummarizedExperiment::colData(se0), SummarizedExperiment::colData(rse)))
```
