---
title: "gap related functions"
author: "Jing Hua Zhao"
output:
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 2
    check_title: false
bibliography: refs.bib
vignette: >
  %\VignetteIndexEntry{gap related functions}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
set.seed(0)
knitr::opts_chunk$set(
  out.extra = 'style="display:block; margin: auto"',
  fig.align = "center",
  fig.path = "gap/",
  collapse = TRUE,
  comment = "#>",
  dev = "png")
```

We describe some functions from R/gap.

## invnormal

The function is widely used in various consortium analyses and defined as follows,
```{r}
invnormal <- function(x) qnorm((rank(x,na.last="keep")-0.5)/sum(!is.na(x)))
```
An example use on data from Poisson distribution is as follows,
```{r}
set.seed(12345)
Ni <- rpois(50, lambda = 4); table(factor(Ni, 0:max(Ni)))
y <- invnormal(Ni)
sd(y)
mean(y)
Ni <- 1:50
y <- invnormal(Ni)
mean(y)
sd(y)
```

## gc.lambda

The definition is as follows,
```{r}
gc.lambda <- function(p) {
  p <- p[!is.na(p)]
  n <- length(p)

  obs <- qchisq(p,1,lower.tail=FALSE)
  exp <- qchisq(1:n/n,1,lower.tail=FALSE)

  lambda <- median(obs)/median(exp)
  return(lambda)
}

# A simplified version is as follows,
# obs <- median(chisq)
# exp <- qchisq(0.5, 1) # 0.4549364
# lambda <- obs/exp
# see also estlambda from GenABEL and qq.chisq from snpStats

# A related function

lambda1000 <- function(lambda, ncases, ncontrols)
  1 + (lambda - 1) * (1 / ncases + 1 / ncontrols)/( 1 / 1000 + 1 / 1000)
```

## METAL.forestplot

The following is the documentation example,
```{r, message = FALSE, fig.height=6, fig.width=12}
require(gap)
require(gap.datasets)
data(OPG)
METAL_forestplot(OPGtbl,OPGall,OPGrsid,width=8.75,height=5)
```

## mhtplot.trunc

Again this is the documentation example,
```r
options(width=120)
require(gap.datasets)
mhtdata <- within(mhtdata, {z=qnorm(p/2, lower.tail=FALSE)})
mhtplot.trunc(mhtdata, chr = "chr", bp = "pos", z = "z", snp = "rsn", y.brk1=10, y.brk2=12, mtext.line=2.5)

# https://portals.broadinstitute.org/collaboration/
# giant/images/0/0f/Meta-analysis_Locke_et_al+UKBiobank_2018.txt.gz
gz <- gzfile("work/Meta-analysis_Locke_et_al+UKBiobank_2018_UPDATED.txt.gz")
BMI <- within(read.delim(gz,as.is=TRUE), {Z <- BETA/SE})
print(subset(BMI[c("CHR","POS","SNP","P")],CHR!=16 & P<=1e-150))
library(Rmpfr)
print(within(subset(BMI, P==0, select=c(CHR,POS,SNP,Z)),
             {P <- format(2*pnorm(mpfr(abs(Z),100),lower.tail=FALSE)); Pvalue <- pvalue(Z); log10P <- -log10p(Z)}))
png("BMI.png", res=300, units="in", width=9, height=6)
par(oma=c(0,0,0,0), mar=c(5,6.5,1,1))
mhtplot.trunc(BMI, chr="CHR", bp="POS", z="Z", snp="SNP",
              suggestiveline=FALSE, genomewideline=-log10(1e-8),
              cex.mtext=1.2, cex.text=1.2,
              annotatelog10P=156, annotateTop = FALSE, highlight=c("rs13021737","rs17817449","rs6567160"),
              mtext.line=3, y.brk1=200, y.brk2=280, cex.axis=1.2, cex.y=1.2, cex=0.5,
              y.ax.space=20,
              col = c("blue4", "skyblue")
)
dev.off()
```
The plot for the GIANT data is shown here, [https://jinghuazhao.github.io/Omics-analysis/BMI/](https://jinghuazhao.github.io/Omics-analysis/BMI/).

## cnvplot and circos.cnvplot

These are a functions to plot distribution of CNVs,

```{r}
library(gap)
cnvplot(cnv)
circos.cnvplot(cnv)
```

## cs

This is a function for credible set.
```r
# zcat METAL/4E.BP1-1.tbl.gz | \
# awk 'NR==1 || ($1==4 && $2 >= 187158034 - 1e6 && $2 < 187158034 + 1e6)' >  4E.BP1.z
tbl <- within(read.delim("4E.BP1.z"),{logp <- logp(Effect/StdErr)})
z <- cs(tbl)
l <- cs(tbl,log_p="logp")
```

## chr_pos_a1_a2 and inv_chr_pos_a1_a2

They are functions to handle SNPid.
```{r}
s <- chr_pos_a1_a2(1,c(123,321),letters[1:2],letters[2:1])
inv_chr_pos_a1_a2(s)
inv_chr_pos_a1_a2("chr1:123-A_B",seps=c(":","-","_"))
```

## snptest_sample

This is a function to output sample file for SNPTEST.
```r
d <- data.frame(ID_1=1,ID_2=1,missing=0,PC1=1,PC2=2,D1=1,P1=10)
snptest_sample(d,C=paste0("PC",1:2),D=paste0("D",1:1),P=paste0("P",1:1))
```
The commands above generates a file named ``snptest.sample.
