---
title: "log(p), log10(p) and p value for a very large z from Normal(0,1)"
author: "Jing Hua Zhao"
output:
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 2
    check_title: false
bibliography: refs.bib
vignette: >
  %\VignetteIndexEntry{plogp}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
set.seed(0)
options(rmarkdown.html_vignette.check_title = FALSE)
knitr::opts_chunk$set(
  out.extra = 'style="display:block; margin: auto"',
  fig.align = "center",
  fig.path = "figures/",
  collapse = TRUE,
  comment = "#>",
  dev = "png")
```

When a Normally distributed association statistic z is very large, its corresponding p value is very small. A genomewide significance is declared at 0.05/1000000=5e-8 with Bonferroni correction assuming 1 million SNPs are tested. This short note describes how to get -log10(p), which can be used in a Q-Q plot and software such as DEPICT. The solution here is generic since z is also the square root of a chi-squared statistic, for instance.

## log(p) and log10(p)
First thing first, here are the answers for log(p) and log10(p) given z,
```r
# log(p) for a standard normal deviate z based on log()
logp <- function(z) log(2)+pnorm(-abs(z), lower.tail=TRUE, log.p=TRUE)

# log10(p) for a standard normal deviate z based on log()
log10p <- function(z) log(2, base=10)+pnorm(-abs(z), lower.tail=TRUE, log.p=TRUE)/log(10)
```
Note `logp()` will be used for functions such as `qnorm()` as in R/gap function `cs()` whereas `log10p()` is more appropriate for Manhattan plot and used in R/gap `sentinels()`.

## Rationale
We start with z=1.96 whose corresponding p value is approximately 0.05.
```r
2*pnorm(-1.96,lower.tail=TRUE)
```
giving an acceptable value 0.04999579, so we proceed to get log10(p)
```r
log10(2)+log10(pnorm(-abs(z),lower.tail=TRUE))
```
leading to the expression above from the fact that log10(X)=log(X)/log(10) since log(),
being the natural log function, ln() -- so log(exp(1)) = 1, in R, works far better on
the numerator of the second term. The use of -abs() just makes sure we are working on
the lower tail of the standard Normal distribution from which our p value is calculated.

## Benchmark
Now we have a stress test,
```r
z <- 20000
-log10p(z)
```
giving -log10(p) = 86858901.

## p, log(p), log10(p) and the multiple precision arithmetic

We would be curious about the p value itself as well, which is furnished with the Rmpfr package
```r
require(Rmpfr)
2*pnorm(mpfr(-abs(z),100),lower.tail=TRUE,log.p=FALSE)
mpfr(log(2),100) + pnorm(mpfr(-abs(z),100),lower.tail=TRUE,log.p=TRUE)
```
giving p = 1.660579603192917090365313727164e-86858901 and -log(p) = -200000010.1292789076808554854177,
respectively. To carry on we have -log10(p) = -log(p)/log(10)=86858901.

To make -log10(p) usable in R we obtain it directly through
```r
as.numeric(-log10(2*pnorm(mpfr(-abs(z),100),lower.tail=TRUE)))
```
which actually yields exactly the same 86858901.

If we go very far to have z=50000. then -log10(p)=542868107 but we have less luck with Rmpfr.

One may wonder the P value in this case, which is 6.6666145952e-542868108 or simply 6.67e-542868108.

The magic function for doing this is defined as follows,
```r
pvalue <- function (z, decimals = 2)
{
    lp <- -log10p(z)
    exponent <- ceiling(lp)
    base <- 10^-(lp - exponent)
    paste0(round(base, decimals), "e", -exponent)
}
```
and it is more appropriate to express p values in scientific format so they can be
handled as follows,
```r
log10pvalue <- function(p=NULL,base=NULL,exponent=NULL)
{
  if(!is.null(p))
  {
    p <- format(p,scientific=TRUE)
    p2 <- strsplit(p,"e")
    base <- as.numeric(lapply(p2,"[",1))
    exponent <- as.numeric(lapply(p2,"[",2))
  } else if(is.null(base) | is.null(exponent)) stop("base and exponent should both be specified")
  log10(base)+exponent
}
```
used as `log10pvalue(p)` when p<=1e-323, or log10pvalue(base=1,exponent=-323) otherwise.

One can also derive logpvalue for natural base (e) similarly.

We end with a quick look-up table
```{r}
require(gap)
v <- data.frame()
for (z in c(5,10,30,50,100,500,1000,2000,3000,5000))
{
  vi <- c(z,pvalue(z),logp(z),log10p(z))
  v <- rbind(v,vi)
}
names(v) <- c("z","P","log(P)","log10(P)")
knitr::kable(v,caption="Table 1. z,P,log(P) and log10(P)")
```

## Application

The `mhtplot.trunc()` function in R/gap accepts three types of arguments:

1. p. P values of association statistics, which could be very small.
2. log10p. log10(P).
3. z. normal statistics that could be very large.

In all three cases, a log10(P) counterpart is obtained internally and to accommodate extreme value,
the y-axis allows for truncation leaving out a given range to highlight the largest.

A plot for the GIANT data is shown here, [https://jinghuazhao.github.io/Omics-analysis/BMI/](https://jinghuazhao.github.io/Omics-analysis/BMI/).
