---
title: "An Overview of pQTLtools"
author: "Jing Hua Zhao"
output:
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 2
    check_title: false
bibliography: refs.bib
vignette: >
  %\VignetteIndexEntry{Overview}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
set.seed(0)
knitr::opts_chunk$set(
  out.extra = 'style="display:block; margin: auto"',
  fig.align = "center",
  fig.path = "figures/",
  collapse = TRUE,
  comment = "#>",
  dev = "png")
```

## cis/trans classification
```{r}
options(width=200)
require(gap)
INF <- Sys.getenv("INF")
f <- file.path(INF,"work","INF1.merge")
clumped <- read.delim(f,as.is=TRUE)
hits <- merge(clumped[c("CHR","POS","MarkerName","prot","log10p")],
              inf1[c("prot","uniprot")],by="prot")
names(hits) <- c("prot","Chr","bp","SNP","log10p","uniprot")
cistrans <- cis.vs.trans.classification(hits,inf1,"uniprot")
cis.vs.trans <- with(cistrans,data)
with(cistrans,table)
with(cistrans,total)
circos.cis.vs.trans.plot(hits=f,inf1,"uniprot")
```
This is used to generate 2d-Manhattan plot
```{r}
INF <- Sys.getenv("INF")
d <- read.csv(file.path(INF,"work","INF1.merge.cis.vs.trans"),as.is=TRUE)
t2d <- mhtplot2d(d)
```
and ideogram
```{r}
INF1_merge <- read.delim(file.path(INF,"work","INF1.merge"))[c("Chrom","Start","End","prot","MarkerName")]
singletons <- with(INF1_merge,End-Start==1)
INF1_merge[singletons,"Start"] <- INF1_merge[singletons,"Start"] - 1e6
INF1_merge[singletons,"End"] <- INF1_merge[singletons,"End"] + 1e6
small <- with(INF1_merge, Start<0)
INF1_merge[small,"Start"] <- 0
cvt <- read.csv(file.path(INF,"work","/INF1.merge.cis.vs.trans"),as.is=TRUE)
INF1_merge_cvt <- merge(INF1_merge,cvt,by.x=c("prot","MarkerName"),by.y=c("prot","SNP"))
ord <- with(INF1_merge_cvt,order(Chr,bp))
INF1_merge_cvt <- INF1_merge_cvt[ord,]

library(biomaRt)
library(regioneR)
mart <- useMart(biomart = "ensembl", dataset = "hsapiens_gene_ensembl")
bm <- with(INF1_merge_cvt, getBM(attributes = c("hgnc_symbol", "chromosome_name", "start_position", "end_position", "band"),
           filters="chromosomal_region", values=paste0(Chr,":",bp,":",bp), mart=mart))
genes <- with(bm,toGRanges(chromosome_name,start_position,end_position,labels=hgnc_symbol))

library(karyoploteR)
attach(INF1_merge_cvt)
sentinels <- toGRanges(Chr,bp-1,bp,labels=p.gene)
cis.regions <- toGRanges(Chr,cis.start,cis.end)
loci <- toGRanges(Chr,Start,End)
panel <- toGRanges(p.chr,p.start,p.end,labels=p.gene)
colors <- c("red","blue")
seqlevelsStyle(sentinels) <- "UCSC"
kp <- plotKaryotype(genome="hg19",chromosomes=levels(seqnames(sentinels)))
kpAddBaseNumbers(kp)
kpPlotRegions(kp,data=loci,r0=0.05,r1=0.15,border="black")
kpPlotMarkers(kp, data=sentinels, labels=p.gene, text.orientation="vertical",
              cex=0.45, y=0.3*seq_along(p.gene)/length(p.gene), srt=30, ignore.chromosome.ends=TRUE,
              adjust.label.position=TRUE, label.color=colors[2-cis], label.dist=0.002,
              cex.axis=3, cex.lab=3)
legend("bottomright", bty="n", pch=c(19,19), col=colors, pt.cex=0.4, legend=c("cis", "trans"), text.col=colors, cex=0.8, horiz=FALSE)
# kpPlotLinks(kp, data=loci, data2=panel, col=colors[2-cis])
detach(INF1_merge_cvt)
```

## Bidirectional Mendelian Randomization

The function `pqtlMR` is derived from @zheng20.

## eQTL Catalogue for colocalization analysis

See example associated with `import_eQTLCatalogue`. A related function is `import_OpenGWAS` used to fetch data from OpenGWAS.

```{r}
library(pQTLtools)
f <- file.path(path.package("pQTLtools"),"eQTL-Catalogue","hg19ToHg38.over.chain")
chain <- rtracklayer::import.chain(f)
invisible(lapply(c("dplyr", "ggplot2", "readr", "coloc", "GenomicRanges","seqminer"), require, character.only = TRUE))
HPC_WORK <- Sys.getenv("HPC_WORK")
gwasvcf::set_bcftools(file.path(HPC_WORK,"bin","bcftools"))
f <- file.path(path.package("pQTLtools"),"eQTL-Catalogue","tabix_ftp_paths.tsv")
tabix_paths <- read.delim(f, sep = "\t", header = TRUE, stringsAsFactors = FALSE) %>% dplyr::as_tibble()
HOME <- Sys.getenv("HOME")
f <- file.path(path.package("pQTLtools"),"eQTL-Catalogue","tabix_ftp_paths_imported.tsv")
imported_tabix_paths <- within(read.delim(f, sep = "\t", header = TRUE, stringsAsFactors = FALSE) %>% dplyr::as_tibble(),
      {ftp_path <- gsub("ftp://ftp.ebi.ac.uk/pub/databases/spot/eQTL/csv/GTEx_V8/ge",
                        paste0(HOME,"/rds/public_databases/GTEx/csv"),ftp_path)})
INF <- Sys.getenv("INF")
M <- 1e6
sentinels <- subset(read.csv(file.path(INF,"work","INF1.merge.cis.vs.trans")),cis)
liftRegion <- function(x,chain)
{
  require(GenomicRanges)
  gr <- with(x,GenomicRanges::GRanges(seqnames=chr,IRanges::IRanges(start,end)))
  seqlevelsStyle(gr) <- "UCSC"
  gr38 <- rtracklayer::liftOver(gr, chain)
  chr <- colnames(table(seqnames(gr38)))
  chr <- gsub("chr","",chr)
  start <- min(unlist(start(gr38)))
  end <- max(unlist(end(gr38)))
  invisible(list(chr=chr,start=start,end=end,region=paste0(chr,":",start,"-",end)))
}

coloc_sumstats <- function(prot,chr,region37)
{
  cat("GWAS sumstats\n")
  gwas_stats <- gwasvcf::query_gwas(file.path(INF,"METAL/gwas2vcf",paste0(prot,".vcf.gz")), chrompos = region37)
  gwas_stats <- gwasvcf::vcf_to_granges(gwas_stats) %>% keepSeqlevels(chr) %>% renameSeqlevels(paste0("chr",chr))
  gwas_stats_hg38 <- rtracklayer::liftOver(gwas_stats, chain) %>%
    unlist() %>%
    renameSeqlevels(chr) %>%
    dplyr::as_tibble() %>%
    dplyr::transmute(chromosome = seqnames, position = start, AF, ES, SE, LP, SS) %>%
    dplyr::mutate(id = paste(chromosome, position, sep = ":")) %>%
    dplyr::mutate(MAF = pmin(AF, 1-AF)) %>%
    dplyr::group_by(id) %>%
    dplyr::mutate(row_count = n()) %>%
    dplyr::ungroup() %>%
    dplyr::filter(row_count == 1)
  ggplot(gwas_stats_hg38, aes(x = position, y = LP)) + geom_point()
  gwas_stats_hg38
}

coloc_c <- function(gwas_stats_hg38,ensGene,region38)
{
  cat("c. GTEx_v8 imported eQTL datasets\n")
  rnaseq_df <- dplyr::filter(imported_tabix_paths, quant_method == "ge") %>% dplyr::mutate(qtl_id = paste(study, qtl_group, sep = "_"))
  ftp_path_list <- setNames(as.list(rnaseq_df$ftp_path), rnaseq_df$qtl_id)
  hdr <- file.path(path.package("pQTLtools"),"eQTL-Catalogue","column_names.GTEx")
  column_names <- names(read.delim(hdr))
  safe_import <- purrr::safely(import_eQTLCatalogue)
  summary_list <- purrr::map(ftp_path_list, ~safe_import(., region38, selected_gene_id = ensGene, column_names))
  result_list <- purrr::map(summary_list, ~.$result)
  result_list <- result_list[!unlist(purrr::map(result_list, is.null))]
  result_filtered <- purrr::map(result_list[lapply(result_list,nrow)!=0], ~dplyr::filter(., !is.na(se)))
  coloc_df_imported <- purrr::map_df(result_filtered, ~run_coloc(., gwas_stats_hg38), .id = "qtl_id")
}

mixed_coloc <- function(prot,chr,ensGene,chain,region37,region38,out)
{
  gwas_stats_hg38 <- coloc_sumstats(prot,chr,region37)
  coloc_df_imported <- coloc_c(gwas_stats_hg38,ensGene,region38)
  if (exists("coloc_df_imported"))
  {
    saveRDS(coloc_df_imported,file=paste0(out,".RDS"))
    dplyr::arrange(coloc_df_imported, -PP.H4.abf)
    ggplot(coloc_df_imported, aes(x = PP.H4.abf)) + geom_histogram()
  }
}

single_run <- function(r)
{
  sentinel <- sentinels[r,]
  isnpid <- within(gap::inv_chr_pos_a1_a2(sentinel[["SNP"]]),
  {
    chr <- gsub("chr","",chr)
    pos <- as.integer(pos)
    start <- pos-M
    if (start<0) start <- 0
    end <- pos+M
  })
  chr <- with(isnpid,chr)
  region37 <- with(isnpid, paste0(chr,":",start,"-",end))
  ensRegion37 <- with(subset(inf1,prot==sentinel[["prot"]]),paste0(chr,":",start,"-",end))
  region38 <- with(liftRegion(isnpid,chain),region)
  ensGene <- subset(inf1,prot==sentinel[["prot"]])[["ensembl_gene_id"]]
  ensRegion38 <- with(liftRegion(subset(inf1,prot==sentinel[["prot"]]),chain),region)
  f <- file.path(INF,"coloc",with(sentinel,paste0(prot,"-",SNP)))
  cat(chr,region37,region38,ensGene,ensRegion37,ensRegion38,"\n")
  mixed_coloc(sentinel[["prot"]],chr,ensGene,chain,region37,region38,f)
}

single_run(1)
```

## Literature on pQTLs

Those as identified by @sun18 and @suhre20 are included as EndNote libraries.

## UniProt IDs

The function `uniprot2ids` converts UniProt IDs to others.

## References
