---
title: "An Overview of pQTLtools"
output:
  bookdown::html_document2:
    toc: true
    toc_float: true
    number_sections: true
fontsize: 11pt
bibliography: '`r system.file("REFERENCES.bib", package="pQTLtools")`'
csl: nature-genetics.csl
pkgdown:
  as_is: false
vignette: >
  %\VignetteIndexEntry{An Overview of pQTLtools}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
set.seed(0)
knitr::opts_chunk$set(
  out.extra = 'style="display:block; margin: auto"',
  fig.align = "center",
  fig.path = "pQTLtools/",
  collapse = TRUE,
  comment = "#>",
  dev = "png")
```

The examples here are based on the SCALLOP work @scallop_inf.

```{r, message=FALSE, warning=FALSE}
pkgs <- c("GenomicRanges", "TwoSampleMR", "biomaRt", "coloc", "dplyr", "gap", "ggplot2", "httr",
          "karyoploteR", "circlize", "knitr", "plotly", "pQTLtools", "rGREAT", "readr", "regioneR",
          "seqminer")
for (p in pkgs) if (length(grep(paste("^package:", p, "$", sep=""), search())) == 0) {
    if (!requireNamespace(p)) warning(paste0("This vignette needs package `", p, "'; please install"))
}
invisible(suppressMessages(lapply(pkgs, require, character.only=TRUE)))
```

# cis/trans classification

## pQTL signals and classification table

```{r ct_data}
options(width=200)
f <- file.path(find.package("pQTLtools"),"tests","INF1.merge")
merged <- read.delim(f,as.is=TRUE)
hits <- merge(merged[c("CHR","POS","MarkerName","prot","log10p")],
              pQTLdata::inf1[c("prot","uniprot")],by="prot") %>%
        dplyr::mutate(log10p=-log10p)
names(hits) <- c("prot","Chr","bp","SNP","log10p","uniprot")
cistrans <- cis.vs.trans.classification(hits,pQTLdata::inf1,"uniprot")
cis.vs.trans <- with(cistrans,data)
kable(with(cistrans,table),caption="cis/trans classification")
with(cistrans,total)
T <- with(cistrans,table)
H <- T[rownames(T)!="total","total"]
merge <- merged[c("Chrom","Start","End","prot","MarkerName")]
merge_cvt <- merge(merge,cis.vs.trans,by.x=c("prot","MarkerName"),by.y=c("prot","SNP"))
ord <- with(merge_cvt,order(Chr,bp))
merge_cvt <- merge_cvt[ord,]
```

## Genomic associations

This is visualised via a circos plot, highlighting the likely causal genes for pQTLs,

```{r pQTLs, fig.cap="Genomic associations", fig.height=9, fig.width=9, messages=FALSE, results="hide", warnings=FALSE}
pQTLs <- transmute(merge_cvt,chr=paste0("chr",Chr),start=bp,end=bp,log10p)
cis.pQTLs <- subset(merge_cvt,cis) %>%
             transmute(chr=paste0("chr",p.chr),start=p.start,end=p.end,gene=p.gene,cols="red")
pQTL_genes <- read.table(file.path(find.package("pQTLtools"),"tests","pQTL_genes.txt"),
                         col.names=c("chr","start","end","gene")) %>%
              mutate(chr=gsub("hs","chr",chr)) %>%
              left_join(cis.pQTLs) %>%
              mutate(cols=ifelse(is.na(cols),"blue",cols))
par(cex=0.7)
gap::circos.mhtplot2(pQTLs,pQTL_genes)
par(cex=1)
```

where the red and blue colours indicate cis/trans classifications.

## Bar chart and circos plot

```{r ct_bar, fig.cap="Bar chart", fig.height=7, fig.width=7}
barplot(table(H),xlab="No. of pQTL regions",ylab="No. of proteins",
        ylim=c(0,25),col="darkgrey",border="black",cex=0.8,cex.axis=2,cex.names=2,las=1)
```

```{r ct_circos, fig.cap="circos plot", fig.height=7, fig.width=7}
circos.cis.vs.trans.plot(hits=f,pQTLdata::inf1,"uniprot")
```

The circos plot is based on target genes (those encoding proteins) and somewhat too busy.

## *SH2B3*

Here we focus on ***SH2B3***.

```{r sh2b3, fig.cap="***SH2B3*** hotspot", fig.height=7, fig.width=7}
  HOTSPOT <- "chr12:111884608_C_T"
  a <- data.frame(chr="chr12",start=111884607,end=111884608,gene="SH2B3")
  b <- filter(cis.vs.trans,SNP==HOTSPOT) %>%
       mutate(p.chr=paste0("chr",p.chr)) %>%
       rename(chr=p.chr,start=p.start,end=p.end,gene=p.gene,cistrans=cis.trans)
  cols <- rep(12,nrow(b))
  cols[b[["cis"]]] <- 10
  labels <- bind_rows(select(b,chr,start,end,gene),a)
  circos.clear()
  circos.par(start.degree=90, track.height=0.1, cell.padding=c(0,0,0,0))
  circos.initializeWithIdeogram(species="hg19", track.height=0.05, ideogram.height=0.06)
  circos.genomicLabels(labels, labels.column=4, cex=1.1, font=4, side="inside")
  circos.genomicLink(bind_rows(a,a,a,a,a,a), b[c("chr","start","end")], col=cols,
                     directional=1, border=10, lwd=2)
```

A more recent implementation is the `qtlClassifier` function. For this example we have,

```{r qtlClassifier, fig.cap="cis/trans classification", fig.height=7, fig.width=7}
options(width=200)
geneSNP <- merge(merged[c("prot","MarkerName")],pQTLdata::inf1[c("prot","gene")],by="prot")[c("gene","MarkerName","prot")]
SNPPos <- merged[c("MarkerName","CHR","POS")]
genePos <- pQTLdata::inf1[c("gene","chr","start","end")]
cvt <- qtlClassifier(geneSNP,SNPPos,genePos,1e6)
kable(head(cvt))
cistrans.check <- merge(cvt[c("gene","MarkerName","Type")],cis.vs.trans[c("p.gene","SNP","cis.trans")],
                        by.x=c("gene","MarkerName"),by.y=c("p.gene","SNP"))
with(cistrans.check,table(Type,cis.trans))
```

## pQTL-gene plot

```{r qtl2dplot, fig.cap="pQTL-gene plot", fig.height=8, fig.width=8}
t2d <- qtl2dplot(cis.vs.trans,xlab="pQTL position",ylab="Gene position")
```

whose result can be also viewed in a 2-d plotly style, [fig2d.html](fig2d.html),

## plotly

These plots are responsive.

```{r qtl2dplotly, fig.cap="2d pQTL-gene plotly", fig.height=8, fig.width=8, plotly=TRUE}
fig2d <- qtl2dplotly(cis.vs.trans,xlab="pQTL position",ylab="Gene position")
htmlwidgets::saveWidget(fig2d,file="fig2d.html")
fig2d
```

and 3-d counterpart, [fig3d.html](fig3d.html),

```{r qtl3dplotly, fig.cap="3d pQTL-gene plotly", fig.height=8, fig.width=8, plotly=TRUE}
fig3d <- qtl3dplotly(cis.vs.trans,zmax=300,qtl.prefix="pQTL:",xlab="pQTL position",ylab="Gene position")
htmlwidgets::saveWidget(fig3d,file="fig3d.html")
fig3d
```

## Karyoplot

As `biomaRt` is not always on, we keep a copy of hgnc.

```r
set_config(config(ssl_verifypeer = 0L))
mart <- useMart(biomart = "ensembl", dataset = "hsapiens_gene_ensembl")
attrs <- c("hgnc_symbol", "chromosome_name", "start_position", "end_position", "band")
hgnc <- vector("character",180)
for(i in 1:180)
{
  v <- with(merge_cvt[i,],paste0(Chr,":",bp,":",bp))
  g <- subset(getBM(attributes = attrs, filters="chromosomal_region", values=v, mart=mart),!is.na(hgnc_symbol))
  hgnc[i] <- paste(g[["hgnc_symbol"]],collapse=";")
  cat(i,g[["hgnc_symbol"]],hgnc[i],"\n")
}
save(hgnc,file="hgnc.rda",compress="xz")
```

We now proceed with

```{r karyoplot, fig.cap="Karyoplot of cis/trans pQTLs", fig.height=7, fig.width=7}
load(file.path(find.package("pQTLtools"),"tests","hgnc.rda"))
merge_cvt <- within(merge_cvt,{
  hgnc <- hgnc
  hgnc[cis] <- p.gene[cis]
})

with(merge_cvt, {
  sentinels <- toGRanges(Chr,bp-1,bp,labels=hgnc)
  cis.regions <- toGRanges(Chr,cis.start,cis.end)
  loci <- toGRanges(Chr,Start,End)
  colors <- c("red","blue")
  seqlevelsStyle(sentinels) <- "UCSC"
  kp <- plotKaryotype(genome="hg19",chromosomes=levels(seqnames(sentinels)))
  kpAddBaseNumbers(kp)
  kpPlotRegions(kp, data=loci,r0=0.05,r1=0.15,border="black")
  kpPlotMarkers(kp, data=sentinels, labels=hgnc, text.orientation="vertical",
                cex=0.5, y=0.3*seq_along(hgnc)/length(hgnc), srt=30,
                ignore.chromosome.ends=TRUE,
                adjust.label.position=TRUE, label.color=colors[2-cis], label.dist=0.002,
                cex.axis=3, cex.lab=3)
  legend("bottomright", bty="n", pch=c(19,19), col=colors, pt.cex=0.4,
         legend=c("cis", "trans"), text.col=colors, cex=0.8, horiz=FALSE)
# panel <- toGRanges(p.chr,p.start,p.end,labels=p.gene)
# kpPlotLinks(kp, data=loci, data2=panel, col=colors[2-cis])
})
```

# Genomic regions enrichment analysis

It is now considerably easier with Genomic Regions Enrichment of Annotations Tool (GREAT).

```{r GREAT, fig.cap="GREAT plots", fig.height=6, fig.width=8, results='hide', warning=FALSE}
post <- function(regions)
{
  job <- submitGreatJob(get(regions), species="hg19", version="3.0.0")
  et <- getEnrichmentTables(job,download_by = 'tsv')
  tb <- do.call('rbind',et)
  write.table(tb,file=paste0(regions,".tsv"),quote=FALSE,row.names=FALSE,sep="\t")
  invisible(list(job=job,tb=tb))
}

M <- 1e+6
merge <- merged %>%
         dplyr::mutate(chr=Chrom, start=POS-M, end=POS+M) %>%
         dplyr::mutate(start=if_else(start<1,1,start)) %>%
         dplyr::select(prot,MarkerName,chr,start,end)
cistrans <- dplyr::select(merge, chr,start,end) %>%
            dplyr::arrange(chr,start,end) %>%
            dplyr::distinct()
# All regions
cistrans.post <- post("cistrans")
job <- with(cistrans.post,job)
plotRegionGeneAssociationGraphs(job)
availableOntologies(job)
# plot of the top term
par(mfcol=c(3,1))
plotRegionGeneAssociationGraphs(job, ontology="GO Molecular Function")
plotRegionGeneAssociationGraphs(job, ontology="GO Biological Process")
plotRegionGeneAssociationGraphs(job, ontology="GO Cellular Component")
# Specific regions
IL12B <- dplyr::filter(merge,prot=="IL.12B") %>% dplyr::select(chr,start,end)
KITLG <- dplyr::filter(merge,prot=="SCF") %>% dplyr::select(chr,start,end)
TNFSF10 <- dplyr::filter(merge,prot=="TRAIL") %>% dplyr::select(chr,start,end)
tb_all <- data.frame()
for (r in c("IL12B","KITLG","TNFSF10"))
{
  r.post <- post(r)
  tb_all <- rbind(tb_all,data.frame(gene=r,with(r.post,tb)))
}
```

The top terms at Binomial p=1e-5 could be extracted as follows,

```{r, echo=FALSE}
options(width=100)
great3 <- filter(tb_all,BinomP<=1e-5) %>%
          dplyr::mutate(ID=gsub(":",": ",ID))
kable(great3,caption="GREAT IL12B-KITLG-TNFSF10 results",digits=3)
great <- read.delim("cistrans.tsv") %>%
         dplyr::filter(BinomBonfP<=1e-8) %>%
         dplyr::mutate(ID=gsub(":",": ",ID))
unlink("IL12B.tsv")
unlink("KITLG.tsv")
unlink("TNFSF10.tsv")
unlink("cistrans.tsv")
```

# eQTL Catalog for colocalization analysis

See example associated with `import_eQTLCatalogue`. A related function is `import_OpenGWAS` used to fetch data from OpenGWAS. The cis-pQTLs and 1e+6 
flanking regions were considered and data are actually fetched from files stored locally. Only the first sentinel was used (r=1).

```{r coloc, fig.cap="Association plot & histogram", fig.height=7, fig.width=8, message=FALSE, warning=FALSE}
liftRegion <- function(x,chain,flanking=1e6)
{
  require(GenomicRanges)
  gr <- with(x,GenomicRanges::GRanges(seqnames=chr,IRanges::IRanges(start,end))+flanking)
  seqlevelsStyle(gr) <- "UCSC"
  gr38 <- rtracklayer::liftOver(gr, chain)
  chr <- gsub("chr","",colnames(table(seqnames(gr38))))
  start <- min(unlist(start(gr38)))
  end <- max(unlist(end(gr38)))
  invisible(list(chr=chr[1],start=start,end=end,region=paste0(chr[1],":",start,"-",end)))
}

sumstats <- function(prot,chr,region37)
{
  cat("GWAS sumstats\n")
  vcf <- file.path(INF,"METAL/gwas2vcf",paste0(prot,".vcf.gz"))
  gwas_stats <- gwasvcf::query_gwas(vcf, chrompos = region37) %>%
                gwasvcf::vcf_to_granges() %>%
                keepSeqlevels(chr) %>%
                renameSeqlevels(paste0("chr",chr))
  gwas_stats_hg38 <- rtracklayer::liftOver(gwas_stats, chain) %>%
    unlist() %>%
    dplyr::as_tibble() %>%
    dplyr::transmute(chromosome = seqnames,
                     position = start, REF, ALT, AF, ES, SE, LP, SS) %>%
    dplyr::mutate(id = paste(chromosome, position, sep = ":")) %>%
    dplyr::mutate(MAF = pmin(AF, 1-AF)) %>%
    dplyr::group_by(id) %>%
    dplyr::mutate(row_count = n()) %>%
    dplyr::ungroup() %>%
    dplyr::filter(row_count == 1) %>%
    dplyr::mutate(chromosome=gsub("chr","",chromosome))
  s <- ggplot2::ggplot(gwas_stats_hg38, aes(x = position, y = LP)) +
       ggplot2::theme_bw() +
       ggplot2::geom_point() +
       ggplot2::ggtitle(with(sentinel,paste0(prot,"-",SNP," association plot")))
  s
  gwas_stats_hg38
}

gtex <- function(gwas_stats_hg38,ensGene,region38)
{
  cat("c. GTEx_v8 imported eQTL datasets\n")
  fp <- file.path(find.package("pQTLtools"),"eQTL-Catalogue","tabix_ftp_paths_gtex.tsv")
  web <- read.delim(fp, stringsAsFactors = FALSE) %>% dplyr::as_tibble()
  local <- within(web %>% dplyr::as_tibble(),
        {
          f <- lapply(strsplit(ftp_path,"/imported/|/ge/"),"[",3);
          ftp_path <- paste0("~/rds/public_databases/GTEx/csv/",f)
        })
  gtex_df <- dplyr::filter(local, quant_method == "ge") %>%
             dplyr::mutate(qtl_id = paste(study, qtl_group, sep = "_"))
  ftp_path_list <- setNames(as.list(gtex_df$ftp_path), gtex_df$qtl_id)
  hdr <- file.path(find.package("pQTLtools"),"eQTL-Catalogue","column_names.GTEx")
  column_names <- names(read.delim(hdr))
  safe_import <- purrr::safely(import_eQTLCatalogue)
  summary_list <- purrr::map(ftp_path_list,
                             ~safe_import(., region38, selected_gene_id = ensGene, column_names))
  result_list <- purrr::map(summary_list, ~.$result)
  result_list <- result_list[!unlist(purrr::map(result_list, is.null))]
  result_filtered <- purrr::map(result_list[lapply(result_list,nrow)!=0],
                                ~dplyr::filter(., !is.na(se)))
  purrr::map_df(result_filtered, ~run_coloc(., gwas_stats_hg38), .id = "qtl_id")
}

gtex_coloc <- function(prot,chr,ensGene,chain,region37,region38,out)
{
  gwas_stats_hg38 <- sumstats(prot,chr,region37)
  df_gtex <- gtex(gwas_stats_hg38,ensGene,region38)
  if (!exists("df_gtex")) return
  saveRDS(df_gtex,file=paste0(out,".RDS"))
  dplyr::arrange(df_gtex, -PP.H4.abf)
  p <- ggplot2::ggplot(df_gtex, aes(x = PP.H4.abf)) +
       ggplot2::theme_bw() +
       geom_histogram() +
       ggtitle(with(sentinel,paste0(prot,"-",SNP," PP4 histogram"))) +
       xlab("PP4") + ylab("Frequency")
  p
}

single_run <- function()
{
  chr <- with(sentinel,Chr)
  ss <- subset(inf1,prot==sentinel[["prot"]])
  ensRegion37 <- with(ss,
                      {
                        start <- start-M
                        if (start<0) start <- 0
                        end <- end+M
                        paste0(chr,":",start,"-",end)
                      })
  ensGene <- ss[["ensembl_gene_id"]]
  ensRegion38 <- with(liftRegion(ss,chain),region)
  cat(chr,ensGene,ensRegion37,ensRegion38,"\n")
  f <- with(sentinel,paste0(prot,"-",SNP))
  gtex_coloc(sentinel[["prot"]],chr,ensGene,chain,ensRegion37,ensRegion38,f)
}

options(width=200)
HOME <- Sys.getenv("HOME")
HPC_WORK <- Sys.getenv("HPC_WORK")
INF <- Sys.getenv("INF")
M <- 1e6
sentinels <- subset(cis.vs.trans,cis)
f <- file.path(find.package("pQTLtools"),"eQTL-Catalogue","hg19ToHg38.over.chain")
chain <- rtracklayer::import.chain(f)
gwasvcf::set_bcftools(file.path(HPC_WORK,"bin","bcftools"))

r <- 1
sentinel <- sentinels[r,]
single_run()
# The results are also loadable as follows.
coloc_df <- readRDS(file.path(find.package("pQTLtools"),"tests","OPG-chr8:120081031_C_T.RDS")) %>%
            dplyr::rename(Tissue=qtl_id, H0=PP.H0.abf,H1=PP.H1.abf,
                          H2=PP.H2.abf,H3=PP.H3.abf,H4=PP.H4.abf) %>%
            mutate(Tissue=gsub("GTEx_V8_","",Tissue),
                   H0=round(H0,2),H1=round(H1,2),H2=round(H2,2),H3=round(H3,2),H4=round(H4,2)) %>%
                   arrange(-H4)
ktitle <- with(sentinel,paste0("Colocalization results for ",prot,"-",SNP))
kable(coloc_df,caption=ktitle)
```

The function `sumstats()` obtained meta-analysis summary statistics (in build 37 and therefore lifted over to build 38) to be used in colocalization 
analysis. The output are saved in the `.RDS` files. Note that `ftp_path` changes from eQTL Catalog to local files.

# Mendelian Randomisation (MR)

## pQTL-based MR

The function `pqtlMR` has an attractive feature that multiple pQTLs
can be used together for conducting MR with a list of outcomes from MR-Base. For generic applications
the `run_TwoSampleMR()` function can be used.

```{r, warning=FALSE}
f <- file.path(system.file(package="pQTLtools"),"tests","Ins.csv")
ivs <- format_data(read.csv(f))
caption4 <- "ABO/LIFR variants and CHD/FEV1"
kable(ivs, caption=paste(caption4,"(instruments)"),digits=3)
ids <- c("ieu-a-7","ebi-a-GCST007432")
pqtlMR(ivs, ids)
result <- read.delim("pQTL-combined-result.txt",header=TRUE)
kable(result,caption=paste(caption4, "(result)"),digits=3)
single <- read.delim("pQTL-combined-single.txt",header=TRUE)
kable(subset(single,!grepl("All",SNP)), caption=paste(caption4, "(single)"),digits=3)
invisible(sapply(c("harmonise","result","single"),
                 function(x) unlink(paste0("pQTL-combined-",x,".txt"))))
```

## Two-sample MR

The documentation example is quoted here,

```{r run_TwoSampleMR, fig.cap="Two-sample MR", fig.height=7, fig.width=7, results='hide', warning=FALSE}
outcomes <- "ebi-a-GCST007432"
prot <- "MMP.10"
type <- "cis"
f <- paste0(prot,"-",type,".mrx")
d <- read.table(file.path(system.file(package="pQTLtools"),"tests",f),
                header=TRUE)
exposure <- format_data(within(d,{P=10^logP}), phenotype_col="prot", snp_col="rsid",
                        chr_col="Chromosome", pos_col="Posistion",
                        effect_allele_col="Allele1", other_allele_col="Allele2",
                        eaf_col="Freq1", beta_col="Effect", se_col="StdErr",
                        pval_col="P", log_pval=FALSE,
                        samplesize_col="N")
clump <- clump_data(exposure)
outcome <- extract_outcome_data(snps=exposure$SNP,outcomes=outcomes)
harmonise <- harmonise_data(clump,outcome)
prefix <- paste(outcomes,prot,type,sep="-")
run_TwoSampleMR(harmonise, mr_plot="pQTLtools", prefix=prefix)
```

The output is contained in individual `.txt` files, together with the scatter, forest, funnel and leave-one-out plots.

```{r, echo=FALSE}
caption5 <- "MMP.10 variants and FEV1"
kable(read.delim(paste0(prefix,"-result.txt"),header=TRUE),caption=paste(caption5, "(result)"),digits=3)
kable(read.delim(paste0(prefix,"-heterogeneity.txt"),header=TRUE),caption=paste(caption5,"(heterogeneity)"),digits=3)
kable(read.delim(paste0(prefix,"-pleiotropy.txt"),header=TRUE),caption=paste(caption5,"(pleiotropy)"),digits=3)
kable(read.delim(paste0(prefix,"-single.txt"),header=TRUE),caption=paste(caption5,"(single)"),digits=3)
kable(read.delim(paste0(prefix,"-loo.txt"),header=TRUE),caption=paste(caption5,"(loo)"),digits=3)
for (x in c("result","heterogeneity","pleiotropy","single","loo")) unlink(paste0(prefix,"-",x,".txt"))
```

## MR using cis, trans and cis+trans (pan) instruments

This is illustrated with IL-12B.

```{r mr, fig.cap="MR with cis, trans and cis+trans variants of IL-12B", fig.height=9}
efo <- read.delim(file.path(find.package("pQTLtools"),"tests","efo.txt"))
d3 <- read.delim(file.path(find.package("pQTLtools"),"tests","IL.12B.txt")) %>%
      dplyr::mutate(MRBASEID=unlist(lapply(strsplit(outcome,"id:"),"[",2)),y=b) %>%
      dplyr::select(-outcome,-method) %>%
      left_join(efo) %>%
      dplyr::arrange(cistrans)
kable(head(d3[c("MRBASEID","trait","cistrans","nsnp","b","se","pval")],29),caption="MR with IL-12B variants",digits=3)
p <- ggplot2::ggplot(d3,aes(y = trait, x = y))+
     ggplot2::theme_bw()+
     ggplot2::geom_point()+
     ggplot2::facet_wrap(~cistrans,ncol=3,scales="free_x")+
     ggplot2::geom_segment(ggplot2::aes(x = b-1.96*se, xend = b+1.96*se, yend = trait))+
     ggplot2::geom_vline(lty=2, ggplot2::aes(xintercept=0), colour = 'red')+
     ggplot2::xlab("Effect size")+
     ggplot2::ylab("")
p
```

## GSMR forest plots

This is based on GSMR results of TNFB and a range of immune-mediated traits,

```{r tnfb, echo=FALSE}
tnfb <- '
              "multiple sclerosis"  0.69058600 0.059270400
    "systemic lupus erythematosus"  0.76687500 0.079000500
          "sclerosing cholangitis"  0.62671500 0.075954700
   "juvenile idiopathic arthritis" -1.17577000 0.160293000
                       "psoriasis"  0.00582586 0.000800016
            "rheumatoid arthritis" -0.00378072 0.000625160
      "inflammatory bowel disease" -0.14334200 0.025272500
          "ankylosing spondylitis" -0.00316852 0.000626225
                  "hypothyroidism" -0.00432054 0.000987324
               "allergic rhinitis"  0.00393075 0.000926002
          "IgA glomerulonephritis" -0.32696600 0.105262000
                   "atopic eczema" -0.00204018 0.000678061
'

tnfb <- as.data.frame(scan(file=textConnection(tnfb),what=list("",0,0))) %>%
        setNames(c("Outcome","Effect","StdErr")) %>%
        mutate(Outcome=gsub("\\b(^[a-z])","\\U\\1",Outcome,perl=TRUE))
kable(tnfb,caption="GSMR results for TNFB and immune-mediated traits",digits=3)
```

```{r mr_forestplot, fig.cap="Forest plots based on MR results on TNFB", fig.align="left", fig.height=8, fig.width=12, results="hide"}

mr_forestplot(tnfb,colgap.forest.left="0.05cm", fontsize=14,
              leftcols=c("studlab"), leftlabs=c("Outcome"),
              plotwidth="5inch", sm="OR", sortvar=tnfb[["Effect"]],
              rightcols=c("effect","ci","pval"), rightlabs=c("OR","95%CI","GSMR P"),
              digits=3, digits.pval=2, scientific.pval=TRUE,
              common=FALSE, random=FALSE, print.I2=FALSE, print.pval.Q=FALSE, print.tau2=FALSE,
              addrow=TRUE, backtransf=TRUE, spacing=1.6, col.inside="black", col.square="black")
```

# Literature on pQTLs

References @sun18 @suhre20 are included as EndNote libraries which is now part of pQTLdata package.

# UniProt IDs

The function `uniprot2ids` converts UniProt IDs to others.

# References
