---
title: Spectrum data analysis
output:
  bookdown::html_document2:
    toc: true
    toc_float: true
fontsize: 11pt
bibliography: '`r system.file("REFERENCES.bib", package="pQTLtools")`'
csl: nature-genetics.csl
link-citations: yes
pkgdown:
  as_is: true
vignette: >
  %\VignetteIndexEntry{Bioconductor notes}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r logo, echo = FALSE, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "spectrum/"
)
```

This article collects notes on peptide/protein analysis, especially with respect to spectrum data.

```{r, message=FALSE, warning=FALSE}
pkgs <- c("Biostrings", "CAMERA", "MSnbase", "MSstats", "Spectra", "mzR", "protViz", "rawrr")
for (p in pkgs) if (length(grep(paste("^package:", p, "$", sep=""), search())) == 0) {
    if (!requireNamespace(p)) warning(paste0("This vignette needs package `", p, "'; please install"))
}
invisible(suppressMessages(lapply(pkgs, require, character.only = TRUE)))
```

# Peptide sequence

Here is an example for PROC_HUMAN, which is handled by the `Biostrings` package,

```r
fasta_file_path <- 'https://rest.uniprot.org/uniprotkb/P04070.fasta'
fasta_sequences <- Biostrings::readAAStringSet(fasta_file_path, format = "fasta")
AA_sequence <- fasta_sequences[[1]]
cat("Sequence:", toString(AA_sequence), "\n")
iso_442688365 <- 'TDGEGALSEPSATVTIEELAAPPPPVLMHHGESSQVLHPGNK'
match_position <- regexpr(iso_442688365, AA_sequence)
match_position
mp <- matchPattern(iso_442688365,AA_sequence)
mp
```

```{r a1bg, fig.cap="peptide association plot",fig.height=6,fig.width=10,messages=FALSE}
load("~/pQTLtools/tests/PROC.rda")
pQTLtools::peptideAssociationPlot(protein,cistrans)
```

# Spectrum data analysis

## Setup

The .raw files can be handled by `rawrr` package nevertheless it requires necessary files,

```r
library(rawrr)
if (isFALSE(rawrr::.checkDllInMonoPath())){
   rawrr::installRawFileReaderDLLs()
}
if (isFALSE(file.exists(rawrr:::.rawrrAssembly()))){
   rawrr::installRawrrExe()
}
```

## Making .raw files ready

Based on a real project, the following is an example of listing/generating multiple .raw from .zip files

```r
# ZWK .raw data
spectra_ZWK <- "~/Caprion/pre_qc_data/spectral_library_ZWK"
raw_files <- list.files(spectra_ZWK, pattern = "\\.raw$", full.names = TRUE)
## collectively
suppressMessages(library(MsBackendRawFileReader))
ZWK <- Spectra::backendInitialize(MsBackendRawFileReader::MsBackendRawFileReader(),
       files = raw_files)
class(ZWK)
methods(class=class(ZWK))
Spectra(ZWK)
spectraData(ZWK)
ZWK
ZWKvars <- ZWK |> Spectra::spectraVariables()
ZWKdata <- ZWK |> Spectra::spectraData()
dim(ZWKdata)
# rows with >=1 non-NA value in the columns with prefix "precursor"
precursor <- apply(ZWKdata[grep("precursor",ZWKvars)], 1, function(x) any(!is.na(x)))
ZWKdata_filtered <- ZWKdata[precursor, ]
save(ZWK,file="~/Caprion/analysis/work/ZWK.rda")

# ZYQ/UDP
library(utils)
spectra <- "~/Caprion/pre_qc_data/spectra"
zip_files <- dir(spectra, recursive = TRUE, full.names=TRUE)
work_dir <- "~/Caprion/analysis/work"
for (zip_file in zip_files) unzip(zip_file, exdir=work_dir)
ZYQ_UDP <- Spectra::backendInitialize(MsBackendRawFileReader::MsBackendRawFileReader(),
           files = dir(work_dir,patt="raw",full.names=TRUE))
class(ZYQ_UDP)
ZYQ_UDP
ZYQ_UDP |> Spectra::spectraVariables()
save(ZYQ_UDP,file="~/Caprion/analysis/work/ZYQ_UDP.rda")
```

## Showcases

We are now ready to expose various facilities in R.

```r
options(width=200)

# mzML
library(mzR)
d <- "/rds/project/rds-MkfvQMuSUxk/interval/caprion_proteomics/spectra"
mz <-  openMSfile(file.path(d,"budp082710c21101xms1.mzML"))
methods(class="mzRpwiz")
spec <- spectra(mz)
class(spec)
length(spec)
lapply(spec,head,3)

# MSnbase
library(MSnbase)
mzXML <- readMSData("szwk901104i19801xms1.mzXML")
mgf <- readMgfData("szwk901104i19801xms1.mgf")
save(mzXML,mgf,file="szwk901104i19801xms1.rda")

extractSpectraData(mzXML)
hasSpectra("szwk901104i19801xms1.mzML,gz")
hasChromatograms("szwk901104i19801xms1.mzML,gz")
plot2d(mzXML,z="peaks.count")
plotDensity(mzXML,z="precursor.mz")

extractSpectraData(mgf)
methods(class="MSpectra")
mz(mgf)
intensity(mgf)
rtime(mgf)
precursorMz(mgf)
precursorCharge(mgf)
precScanNum(mgf)
precursorIntensity(mgf)
acquisitionNum(mgf)
scanIndex(mgf)
peaksCount(mgf)
msLevel(mgf)
tic(mgf)
ionCount(mgf)
collisionEnergy(mgf)
fromFile(mgf)
polarity(mgf)
smoothed(mgf)
centroided(mgf)
isCentroided(mgf)
writeMgfData(mgf, con = "spectra.mgf", COM = NULL, TITLE = NULL)
removePeaks(mgf, t, msLevel., ...)
filterMsLevel(mgf, msLevel=2)
as.ExpressionSet(mgf)

# This turned to be really slow!
sp_list <- lapply(seq_along(mgf), function(i) {
  intensity_i <- intensity(mgf)[[i]]
  mz_i <- mz(mgf)[[i]]
  centroided_i <- centroided(mgf)[[i]]
  return(new("Spectrum1", intensity = intensity_i, mz = mz_i, centroided = centroided_i))
})
sp1 <- do.call(rbind, sp_list)
# only the first one is more manageable
sp1 <- new("Spectrum1",intensity=intensity(mgf)[[1]],mz=mz(mgf)[[1]],centroided=centroided(mgf)[[1]])
sp2 <- pickPeaks(sp1)
intensity(sp2)
plot(mz(sp1),intensity(sp1),type="h")
## Without m/z refinement
points(mz(sp2), intensity(sp2), col = "darkgrey")
## Using k = 1, closest signals
sp3 <- pickPeaks(sp1, refineMz = "kNeighbors", k = 1)
points(mz(sp3), intensity(sp3), col = "green", type = "h")
## Using descendPeak requiring at least 50% or the centroid's intensity
sp4 <- pickPeaks(sp1, refineMz = "descendPeak", signalPercentage = 50)
points(mz(sp4), intensity(sp4), col = "red", type = "h")

# CAMERA
library(CAMERA)
xs   <- xcmsSet(c("szwk901104i19801xms1.mzML"), method="centWave", ppm=30, peakwidth=c(5,10))
an   <- xsAnnotate(xs)
an   <- groupFWHM(an)
#For one group
peaklist <- getpspectra(an, 1)
#For two groups
peaklist <- getpspectra(an, c(1,2))

# Spectra
suppressMessages(library(Spectra))
f <- "szwk901104i19801xms1.mzML,gz"
sp <- Spectra(f)
head(sp)
table(sp$msLevel)
d <- computeMzDeltas(sp[1:1000])
plotMzDelta(d)

# protViz
library(protViz)
protViz::fragmentIon("TFVLNFIK")
esd <- extractSpectraData(mgf)
op <- par(mfrow=c(2,1))
ms <- function(i) with(esd[i,],list(title=TITLE,rtinseconds=RTINSECONDS,pepmass=PEPMASS,charge=CHARGE,
                                    mZ=mz(mgf[[i]]),intensity=intensity(mgf[[i]])))
peakplot("TAFDEAIAELDTLNEESYK", ms(1))
peakplot("TAFDEAIAELDTLSEESYK", ms(2))
par(op)
load("~/Caprion/pilot/ZWK.rda")
peptides <- subset(mapping_ZWK,Protein=="PROC_HUMAN")[["Modified.Peptide.Sequence"]] |> unique()
pim <- parentIonMass(peptides)
fi <- fragmentIon(peptides)
df <- as.data.frame(fi)
pdf("~/Caprion/analysis/peptide/PROC/PROC-fragmention.pdf",width=10,height=12)
op <- par(mfrow=c(3,1))
for (i in 1:length(peptides)){
    plot(0, 0,
    xlab='m/Z',
    ylab='',
    xlim=range(c(fi[[i]]$b,fi[[i]]$y)),
    ylim=c(0,1),
    type='n',
    axes=FALSE,
    sub=paste(peptides[i], "/", pim[i], "Da"));
    box()
    axis(1, fi[[i]]$b, round(fi[[i]]$b,1), las=2)
    axis(1, fi[[i]]$y, round(fi[[i]]$y,1), las=2)

    pepSeq<-strsplit(peptides[i], "")
    axis(3,fi[[i]]$b, paste("b", row.names(fi[[i]]),sep=''),las=2)
    axis(3,fi[[i]]$y, paste("y", row.names(fi[[i]]),sep=''),las=2)

    text(fi[[i]]$b, rep(0.3, nchar(peptides[i])),
    pepSeq[[1]],pos=3,cex=4, lwd=4, col="#aaaaaaaa")

    abline(v=fi[[i]]$b, col='red')
    abline(v=fi[[i]]$y, col='blue',lwd=2)
}
par(op)
dev.off()

# MSstats
library(MSstats)
f <- "szwk901104i19801xms1.mzML,gz"
x <- mzR::openMSfile(f, backend = "pwiz")
x
nChrom(x)
head(tic(x))
head(chromatogram(x, 1L)) ## same as tic(x)
str(chromatogram(x))
p <- mzR::peaks(x)
head(peaks(x, scan=4))

head(SRMRawData)
QuantData <- dataProcess(SRMRawData, use_log_file = FALSE)
head(QuantData$FeatureLevelData)

quant <- dataProcess(SRMRawData,
                      normalization = "equalizeMedians",
                      summaryMethod = "TMP",
                      censoredInt = "NA",
                      MBimpute = TRUE,
                      maxQuantileforCensored = 0.999)
head(quant)
```

# Bioconductor/CRAN packages

Package | Description
--------|------------
**Bioconductor** |
Biostrings | Efficient manipulation of biological strings
CAMERA | Collection of annotation related methods
MSnbase | Base Functions and Classes for Mass Spectrometry and Proteomics
MSstats | Protein Significance Analysis in DDA, SRM and DIA Proteomics
Spectra | Spectra Infrastructure for Mass Spectrometry
mzR | parser for netCDF, mzXML and mzML and mzIdentML files
rawrr | Direct Access to Orbitrap Data and Beyond
**CRAN** |
protViz | Foreach Parallel Adaptor for the 'parallel' Package

# References
