---
output: github_document
---

```{r logo, echo = FALSE, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "figures/"
)
```

## Transcript databases

```r
options(width=200)

suppressMessages(library(AnnotationDbi))
suppressMessages(library(org.Hs.eg.db))
columns(org.Hs.eg.db)
keyref <- keys(org.Hs.eg.db, keytype="ENTREZID")
symbol_uniprot <- select(org.Hs.eg.db,keys=keyref,columns = c("SYMBOL","UNIPROT"))
subset(symbol_uniprot,SYMBOL=="MC4R")

suppressMessages(library(EnsDb.Hsapiens.v86))
x <- EnsDb.Hsapiens.v86
listColumns(x, "protein", skip.keys=TRUE)
listGenebiotypes(x)
listTxbiotypes(x)
listTables(x)
metadata(x)
organism(x)
returnFilterColumns(x)
seqinfo(x)
seqlevels(x)
updateEnsDb(x)

suppressMessages(library(ensembldb))
genes(x, columns=c("gene_name"), filter=list(SeqNameFilter("X"), GeneBiotypeFilter("protein_coding")))
transcripts(x, columns=listColumns(x, "tx"), filter = AnnotationFilterList(), order.type = "asc", return.type = "GRanges")

library(RMariaDB)
library(GenomicFeatures)
txdbEnsemblGRCh38 <- makeTxDbFromEnsembl(organism="Homo sapiens", release=98)
txdb <- as.list(txdbEnsemblGRCh38)
lapply(txdb,head)

library(TxDb.Hsapiens.UCSC.hg38.knownGene)
txdb <- TxDb.Hsapiens.UCSC.hg38.knownGene

library(INSPEcT)
liverExprs <- quantifyExpressionsFromBWs(txdb = txdb,BWfiles=,experimentalDesign=)
```

## Pathway and enrichment analysis

```r
library(graphite)
reactome <- pathways("hsapiens", "reactome")
kegg <- pathways("hsapiens","kegg")
pharmgkb <- pathways("hsapiens","pharmgkb")
nodes(kegg)
suppressMessages(library(plyr))
kegg_t2g <- ldply(lapply(kegg, nodes), data.frame)
names(kegg_t2g) <- c("gs_name", "gene_symbol")
suppress(library(clusterProfiler))
eKEGG <- enricher(gene = , TERM2GENE = kegg_t2g,
                  universe = ,
                  pAdjustMethod = "BH",
                  pvalueCutoff = 0.1, qvalueCutoff = 0.05,
                  minGSSize = 10, maxGSSize = 500)

```

## Differential expression

```{r DESeq2}
suppressMessages(library(DESeq2))
ex <- makeExampleDESeqDataSet(m=4)
dds <- DESeq(ex)
res <- results(dds, contrast=c("condition","B","A"))
rld <- rlogTransformation(ex, blind=TRUE)
dat <- plotPCA(rld, intgroup=c("condition"),returnData=TRUE)
percentVar <- round(100 * attr(dat,"percentVar"))
suppressMessages(library(ggplot2))
ggplot(dat, aes(PC1, PC2, color=condition, shape=condition)) +
geom_point(size=3) +
xlab(paste0("PC1:",percentVar[1],"% variance")) +
ylab(paste0("PC2:",percentVar[2],"% variance"))
ex$condition <- relevel(ex$condition, ref="B")
dds2 <- DESeq(dds)
res <- results(dds2)
write.csv(as.data.frame(res),file="A_vs_B.csv")
```

## Gene co-expression and network analysis

A simple network is furnished with the `GeneNet` documentation example,

```{r GeneNet}
library("GeneNet")

## A random network with 40 nodes 
# it contains 780=40*39/2 edges of which 5 percent (=39) are non-zero
true.pcor <- ggm.simulate.pcor(40)
  
# A data set with 40 observations
m.sim <- ggm.simulate.data(40, true.pcor)

# A simple estimate of partial correlations
estimated.pcor <- cor2pcor( cor(m.sim) )

# A comparison of estimated and true values
sum((true.pcor-estimated.pcor)^2)

# A slightly better estimate ...
estimated.pcor.2 <- ggm.estimate.pcor(m.sim)
sum((true.pcor-estimated.pcor.2)^2)

## ecoli data 
data(ecoli)

# partial correlation matrix 
inferred.pcor <- ggm.estimate.pcor(ecoli)

# p-values, q-values and posterior probabilities for each potential edge 
test.results <- network.test.edges(inferred.pcor)

# best 20 edges (strongest correlation)
test.results[1:20,]

# network containing edges with prob > 0.9 (i.e. local fdr < 0.1)
net <- extract.network(test.results, cutoff.ggm=0.9)
net

# significant based on FDR cutoff Q=0.05?
num.significant.1 <- sum(test.results$qval <= 0.05)
test.results[1:num.significant.1,]

# significant based on "local fdr" cutoff (prob > 0.9)?
num.significant.2 <- sum(test.results$prob > 0.9)
test.results[test.results$prob > 0.9,]

# parameters of the mixture distribution used to compute p-values etc.
c <- fdrtool(sm2vec(inferred.pcor), statistic="correlation")
c$param

## A random network with 20 nodes and 10 percent (=19) edges
true.pcor <- ggm.simulate.pcor(20, 0.1)

# convert to edge list
test.results <- ggm.list.edges(true.pcor)[1:19,]

# graphviz
network.make.dot(filename="test.dot", test.results, nlab, main = "A graph")
system("fdp -T svg -o test.svg test.dot")

# Rgraphviz
nlab <- LETTERS[1:20]
gr <- network.make.graph( test.results, nlab)
gr
num.nodes(gr)
edge.info(gr)
gr2 <- network.make.graph( test.results, nlab, drop.singles=TRUE)
gr2
num.nodes(gr2)
edge.info(gr2)

# plot network
library("Rgraphviz")
plot(gr, "fdp")
plot(gr2, "fdp")
```

and a more involved version

```r
set.seed(123454321)
m <- matrix(runif(2500),50)
r <- cor(m)
g <- as.matrix(r>=0.7)+0
f1 <- Heatmap(r)
f2 <- Heatmap(g)
f <- f1+f2
draw(f) # f2 is somewhat twisted
suppressMessages(library(WGCNA))
pwr <- c(1:10, seq(from = 12, to=30, by=2))
sft <- pickSoftThreshold(dat, powerVector = pwr, verbose = 5)
meg <- moduleEigengenes(t(tpm), color=Colors, softPower=6)
TOM <- TOMsimilarity(adjMatrix)
Tree <- hclust(as.dist(1-TOM), method = "average")
plotDendroAndColors(Tree, colors, "Dynamic Tree Cut", dendroLabels = FALSE, hang = 0.03, addGuide = TRUE, guideHang = 0.05)
```

## Meta-data

```r
suppressMessages(library(recount3))
hs <- available_projects()
dim(subset(hs,file_source=="gtex"))
annotation_options("human")
blood_rse <- create_rse(subset(hs,project=="BLOOD"))
metadata(blood_rse)
dim(blood_rse)
rowRanges(blood_rse)
colnames(colData(blood_rse))
expand_sra_attributes(blood_rse)
```

---

## A list of Bioconductor/CRAN packages

Package | Description
--------|------------
**Bioconductor** |
AnnotationDbi | AnnotationDb objects and their progeny, methods etc.
org.Hs.eg.db | Conversion of Entrez ID -- gene symbols
EnsDb.Hsapiens.v86 | Exposes an annotation databases generated from Ensembl
ensembldb | Retrieve annotation data from an Ensembl based package
TxDb.Hsapiens.UCSC.hg38.knownGene | Annotation of the human genome
INSPEcT | Quantification of the intronic and exonic gene features and the post-transcriptional regulation analysis
graphite | GRAPH Interaction from pathway topological environment
clusterProfiler | Functional profiles for genes and gene clusters
DESSeq2 | Differential gene expression analysis based on the negative binomial distribution
edgeR | Empirical analysis of digital gene expression
WGCNA | Weighted correlation network analysis
ComplexHeatmap | Make complex heatmaps
recount3 | Interface to uniformly processed RNA-seq data
Pi | Priority index, leveraging genetic evidence to prioritise drug targets at the gene and pathway level
Rgraphiz | Interfaces R with the AT&T graphviz library for plotting R graph objects from the graph package
**CRAN** |
GeneNet | Modeling and Inferring Gene Networks
ggplot2 | Data Visualisations Using the grammar of graphics
pheatmap | results visualisation
plyr | Splitting, applying and combining data
