---
title: Bioconductor Notes
output:
  html_document:
    mathjax:  default
    fig_caption:  true
    toc: true
    section_numbering: true
bibliography: bioc.bib
link-citations: yes
csl: nature-genetics.csl
vignette: >
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteIndexEntry{Bioconductor Applications}
  %\VignetteEncoding{UTF-8}
---

```{r logo, echo = FALSE, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "bioconductor/"
)
```

This article collects notes on Bioconductor packages, made available here to faciliate their use and extensions.

## Normalization

### ComBat

This is the documentation example, based on Bioconductor 3.14.

```{r ComBat, messages=FALSE}
requireNamespace("sva")
suppressMessages(require("bladderbatch"))
data(bladderdata)
dat <- bladderEset[1:50,]

pheno = pData(dat)
edata = exprs(dat)
batch = pheno$batch
mod = model.matrix(~as.factor(cancer), data=pheno)

# 1. parametric adjustment
combat_edata1 = sva::ComBat(dat=edata, batch=batch, mod=NULL, par.prior=TRUE, prior.plots=TRUE)

# 2. non-parametric adjustment, mean-only version
combat_edata2 = sva::ComBat(dat=edata, batch=batch, mod=NULL, par.prior=FALSE, mean.only=TRUE)

# 3. reference-batch version, with covariates
combat_edata3 = sva::ComBat(dat=edata, batch=batch, mod=mod, par.prior=TRUE, ref.batch=3, prior.plots=TRUE)
```

### quantro

Again, this is adapted from the package vignette. noting that `FlowSorted.DLPFC.450k` instead of `FlowSorted` is used.

```{r quantro, messages=FALSE}
suppressMessages(requireNamespace("FlowSorted.DLPFC.450k"))
suppressMessages(require("FlowSorted.DLPFC.450k"))
p <- minfi::getBeta(FlowSorted.DLPFC.450k,offset=100)
pd <- Biobase::pData(FlowSorted.DLPFC.450k)
requireNamespace("quantro")
quantro::matdensity(p, groupFactor = pd$CellType, xlab = " ", ylab = "density",
                    main = "Beta Values", brewer.n = 8, brewer.name = "Dark2")
legend('top', c("NeuN_neg", "NeuN_pos"), col = c(1, 2), lty = 1, lwd = 3)
quantro::matboxplot(p, groupFactor = pd$CellType, xaxt = "n", main = "Beta Values")
qtest <- quantro::quantro(object = p, groupFactor = pd$CellType)
# commented since `R CMD check` returns error with multithreads
# requireNamespace("doParallel")
# registerDoParallel(cores=10)
# qtestPerm <- quantro::quantro(p, groupFactor = pd$CellType, B = 1000)
# quantro::qtestPerm
# quantro::quantroPlot(qtestPerm)
```

## Differential expression

```{r DESeq2, messages=FALSE}
requireNamespace("DESeq2")
ex <-  DESeq2::makeExampleDESeqDataSet(m=4)
dds <-  DESeq2::DESeq(ex)
res <-  DESeq2::results(dds, contrast=c("condition","B","A"))
rld <- DESeq2::rlogTransformation(ex, blind=TRUE)
dat <- DESeq2::plotPCA(rld, intgroup=c("condition"),returnData=TRUE)
percentVar <- round(100 * attr(dat,"percentVar"))
requireNamespace("ggplot2")
ggplot2::ggplot(dat, ggplot2::aes(PC1, PC2, color=condition, shape=condition)) +
ggplot2::geom_point(size=3) +
ggplot2::xlab(paste0("PC1:",percentVar[1],"% variance")) +
ggplot2::ylab(paste0("PC2:",percentVar[2],"% variance"))
ex$condition <- stats::relevel(ex$condition, ref="B")
dds2 <- DESeq2::DESeq(dds)
res <- DESeq2::results(dds2)
knitr::kable(head(as.data.frame(res)))
```

## Gene co-expression and network analysis

A simple network is furnished with the `GeneNet` documentation example,

```{r GeneNet, messages=FALSE}
requireNamespace("GeneNet")
require("GeneNet")
## A random network with 40 nodes 
# it contains 780=40*39/2 edges of which 5 percent (=39) are non-zero
true.pcor <- GeneNet::ggm.simulate.pcor(40)
  
# A data set with 40 observations
m.sim <- GeneNet::ggm.simulate.data(40, true.pcor)

# A simple estimate of partial correlations
estimated.pcor <- corpcor::cor2pcor( cor(m.sim) )

# A comparison of estimated and true values
sum((true.pcor-estimated.pcor)^2)

# A slightly better estimate ...
estimated.pcor.2 <- GeneNet::ggm.estimate.pcor(m.sim)
sum((true.pcor-estimated.pcor.2)^2)

## ecoli data 
data(ecoli)

# partial correlation matrix 
inferred.pcor <- GeneNet::ggm.estimate.pcor(ecoli)

# p-values, q-values and posterior probabilities for each potential edge 
test.results <- GeneNet::network.test.edges(inferred.pcor)

# best 20 edges (strongest correlation)
test.results[1:20,]

# network containing edges with prob > 0.9 (i.e. local fdr < 0.1)
net <- GeneNet::extract.network(test.results, cutoff.ggm=0.9)
net

# significant based on FDR cutoff Q=0.05?
num.significant.1 <- sum(test.results$qval <= 0.05)
test.results[1:num.significant.1,]

# significant based on "local fdr" cutoff (prob > 0.9)?
num.significant.2 <- sum(test.results$prob > 0.9)
test.results[test.results$prob > 0.9,]

# parameters of the mixture distribution used to compute p-values etc.
c <- fdrtool::fdrtool(sm2vec(inferred.pcor), statistic="correlation")
c$param

## A random network with 20 nodes and 10 percent (=19) edges
true.pcor <- GeneNet::ggm.simulate.pcor(20, 0.1)

# convert to edge list
test.results <- GeneNet::ggm.list.edges(true.pcor)[1:19,]
nlab <- LETTERS[1:20]

# graphviz
# GeneNet::network.make.dot(filename="test.dot", test.results, nlab, main = "A graph")
# system("fdp -T svg -o test.svg test.dot")

# Rgraphviz
gr <- GeneNet::network.make.graph( test.results, nlab)
gr
GeneNet::num.nodes(gr)
GeneNet::edge.info(gr)
gr2 <- GeneNet::network.make.graph( test.results, nlab, drop.singles=TRUE)
gr2
GeneNet::num.nodes(gr2)
GeneNet::edge.info(gr2)

# plot network
requireNamespace("Rgraphviz")
Rgraphviz::plot(gr, "fdp")
Rgraphviz::plot(gr2, "fdp")
```

A side-by-side heatmaps

```{r heatmap, messages=FALSE}
requireNamespace("ComplexHeatmap")
set.seed(123454321)
m <- matrix(runif(2500),50)
r <- cor(m)
g <- as.matrix(r>=0.7)+0
f1 <- ComplexHeatmap::Heatmap(r)
f2 <- ComplexHeatmap::Heatmap(g)
f <- f1+f2
ComplexHeatmap::draw(f)

requireNamespace("RColorBrewer")
requireNamespace("heatmaply")
df <- heatmaply::normalize(mtcars)
heatmaply::heatmaply(df,k_col=5,k_row=5,colors = colorRampPalette(RColorBrewer::brewer.pal(3, "RdBu"))(256))
```

and a module analysis with WGCNA,

```r
suppressMessages(require(WGCNA))
suppressMessages(require(cluster))

pwr <- c(1:10, seq(from=12, to=30, by=2))
sft <- WGCNA::pickSoftThreshold(dat, powerVector=pwr, verbose=5)
ADJ <- abs(cor(dat, method="pearson", use="pairwise.complete.obs"))^6
dissADJ <- 1-ADJ
dissTOM <- WGCNA::TOMdist(ADJ)
TOM <- WGCNA::TOMsimilarityFromExpr(dat)
Tree <- stats::hclust(stats::as.dist(1-TOM), method="average")
for(j in pwr)
{
  pam_name <- paste0("pam",j)
  assign(pam_name, cluster::pam(as.dist(dissADJ),j))
  pamTOM_name <- paste0("pamTOM",j)
  assign(pamTOM_name,cluster::pam(as.dist(dissTOM),j))
  tc <- table(get(pam_name)$clustering,get(pamTOM_name)$clustering)
  print(tc)
  print(diag(tc))
}
colorStaticTOM <- as.character(WGCNA::cutreeStaticColor(Tree,cutHeight=.99,minSize=5))
colorDynamicTOM <- WGCNA::labels2colors(WGCNA::cutreeDynamic(Tree,method="tree",minClusterSize=5))
Colors <- data.frame(pamTOM6$clustering,colorStaticTOM,colorDynamicTOM)
WGCNA::plotDendroAndColors(Tree, Colors, dendroLabels=FALSE, hang=0.03, addGuide=TRUE, guideHang=0.05)
meg <- WGCNA::moduleEigengenes(dat, color=1:ncol(dat), softPower=6)
```

## Meta-data

```{r meta-data, messages=FALSE}
suppressMessages(requireNamespace("recount3"))
requireNamespace("SummarizedExperiment")
requireNamespace("S4Vectors")
hs <- recount3::available_projects()
dim(subset(hs,file_source=="gtex"))
recount3::annotation_options("human")
blood_rse <- recount3::create_rse(subset(hs,project=="BLOOD"))
S4Vectors::metadata(blood_rse)
SummarizedExperiment::rowRanges(blood_rse)
colnames(SummarizedExperiment::colData(blood_rse))[1:20]
recount3::expand_sra_attributes(blood_rse)
```

## Pathway and enrichment analysis

```{r pathway-enrichment, messages=FALSE}
requireNamespace("graph")
requireNamespace("graphite")
reactome <- graphite::pathways("hsapiens", "reactome")
kegg <- graphite::pathways("hsapiens","kegg")
pharmgkb <- graphite::pathways("hsapiens","pharmgkb")
graph::nodes(kegg[[21]])
requireNamespace("plyr")
kegg_t2g <- plyr::ldply(lapply(kegg, graph::nodes), data.frame)
names(kegg_t2g) <- c("gs_name", "gene_symbol")
requireNamespace("clusterProfiler")
VEGF <- subset(kegg_t2g,gs_name=="VEGF signaling pathway")[[2]]
eKEGG <- clusterProfiler::enricher(gene=VEGF, TERM2GENE = kegg_t2g,
                                   universe=,
                                   pAdjustMethod = "BH",
                                   pvalueCutoff = 0.1, qvalueCutoff = 0.05,
                                   minGSSize = 10, maxGSSize = 500)

```

## Transcript databases

An overview of annotation is available @carlson16.

```{r ts-data, messages=FALSE}
options(width=200)

requireNamespace("AnnotationDbi")
require("org.Hs.eg.db")
AnnotationDbi::columns(org.Hs.eg.db)
keyref <- AnnotationDbi::keys(org.Hs.eg.db, keytype="ENTREZID")
symbol_uniprot <- AnnotationDbi::select(org.Hs.eg.db,keys=keyref,columns = c("SYMBOL","UNIPROT"))
subset(symbol_uniprot,SYMBOL=="MC4R")

require("EnsDb.Hsapiens.v86")
x <- EnsDb.Hsapiens.v86
requireNamespace("ensembldb")
ensembldb::listColumns(x, "protein", skip.keys=TRUE)
ensembldb::listGenebiotypes(x)
ensembldb::listTxbiotypes(x)
ensembldb::listTables(x)
ensembldb::metadata(x)
ensembldb::organism(x)
ensembldb::returnFilterColumns(x)
ensembldb::seqinfo(x)
ensembldb::seqlevels(x)
ensembldb::updateEnsDb(x)

ensembldb::genes(x, columns=c("gene_name"), filter=list(SeqNameFilter("X"), GeneBiotypeFilter("protein_coding")))
ensembldb::transcripts(x, columns=listColumns(x, "tx"), filter = AnnotationFilterList(), order.type = "asc", return.type = "GRanges")

requireNamespace("RMariaDB")
requireNamespace("GenomicFeatures")
txdbEnsemblGRCh38 <- GenomicFeatures::makeTxDbFromEnsembl(organism="Homo sapiens", release=98)
txdb <- as.list(txdbEnsemblGRCh38)
lapply(txdb,head)

require("TxDb.Hsapiens.UCSC.hg38.knownGene")
txdb <- TxDb.Hsapiens.UCSC.hg38.knownGene

# requireNamespace("INSPEcT")
# liverExprs <- INSPEcT::quantifyExpressionsFromBWs(txdb = txdb,BWfiles=,experimentalDesign=)
```

---

## A list of Bioconductor/CRAN packages

Package | Description
--------|------------
**Bioconductor** |
AnnotationDbi | AnnotationDb objects and their progeny, methods etc.
Biobase | Base functions for Bioconductor
org.Hs.eg.db | Conversion of Entrez ID -- gene symbols
EnsDb.Hsapiens.v86 | Exposes an annotation databases generated from Ensembl
ensembldb | Retrieve annotation data from an Ensembl based package
TxDb.Hsapiens.UCSC.hg38.knownGene | Annotation of the human genome
INSPEcT | Quantification of the intronic and exonic gene features and the post-transcriptional regulation analysis
graphite | GRAPH Interaction from pathway topological environment
clusterProfiler | Functional profiles for genes and gene clusters
DESSeq2 | Differential gene expression analysis based on the negative binomial distribution
edgeR | Empirical analysis of digital gene expression
WGCNA | Weighted correlation network analysis
ComplexHeatmap | Make complex heatmaps
recount3 | Interface to uniformly processed RNA-seq data
Pi | Priority index, leveraging genetic evidence to prioritise drug targets at the gene and pathway level
quantro | A test for when to use quantile normalization
FlowSorted.DLPFC.450k | Illumina HumanMethylation data on sorted frontal cortex cell populations
Rgraphiz | Interfaces R with the AT&T graphviz library for plotting R graph objects from the graph package
sva | Surrogate Variable Analysis
doParallel | Foreach Parallel Adaptor for the 'parallel' Package
**CRAN** |
GeneNet | Modeling and Inferring Gene Networks
RColorBrewer | ColorBrewer Palettes
ggplot2 | Data Visualisations Using the grammar of graphics
heatmaply | Interactive Cluster Heat Maps Using `plotly` and `ggplot2`
pheatmap | results visualisation
plyr | Splitting, applying and combining data

## References
